# Optimization

```{r}
library(ggplot2)

theme_set(
  theme_minimal(base_size = 14)
)
```

Optimization plays an important role in statistical computing. Many statistical estimators can be written as

$$
\hat{\theta}
=
\arg\min_{\theta} Q(\theta)
\qquad\text{or}\qquad
\hat{\theta}
=
\arg\max_{\theta} \ell(\theta),
$$

where $Q(\theta)$ is an objective function and $\ell(\theta)$ may be a likelihood.

For simple problems, a closed-form solution may exist. For many statistical models, however, the solution must be obtained numerically.

Examples include maximum likelihood estimation, nonlinear regression, mixture models, machine learning, smoothing, and regularization.

The central idea is

$$
\boxed{
\text{Statistical Problem}
\Longleftrightarrow
\text{Optimization Problem}
\Longleftrightarrow
\text{Numerical Algorithm}
\Longleftrightarrow
\text{R Implementation}
}.
$$

## Why Numerical Optimization?

Consider

$$
X_1,\ldots,X_n
\stackrel{iid}{\sim}
N(\mu,1).
$$

From mathematical statistics,

$$
\hat{\mu}_{MLE}
=
\bar X.
$$

We already know the closed-form solution. However, we can also obtain the estimator computationally by minimizing the negative log-likelihood.

```{r}
set.seed(8670)

x <- rnorm(
  n = 50,
  mean = 2,
  sd = 1
)

nll <- function(mu) {
  -sum(
    dnorm(
      x,
      mean = mu,
      sd = 1,
      log = TRUE
    )
  )
}

fit <- optimize(
  nll,
  interval = c(-1, 5)
)

c(
  numerical = fit$minimum,
  theoretical = mean(x)
)
```

### Visualizing the Objective Function

Rather than treating an optimizer as a black box, we can evaluate the objective function over a grid.

```{r}
mu_grid <- seq(
  -1,
  5,
  length.out = 400
)

nll_data <- data.frame(
  mu = mu_grid,
  nll = vapply(
    mu_grid,
    nll,
    numeric(1)
  )
)

ggplot(
  nll_data,
  aes(
    x = mu,
    y = nll
  )
) +
  geom_line(linewidth = 1) +
  geom_vline(
    xintercept = mean(x),
    linetype = "dashed"
  ) +
  labs(
    x = expression(mu),
    y = "Negative log-likelihood",
    title = "MLE as Numerical Optimization"
  )
```

::: {.callout-question title="In-Class Question"}

Why is numerical optimization useful here if we already know that

$$
\hat{\mu}_{MLE}=\bar X?
$$

:::

The closed-form solution gives us a way to **validate the numerical algorithm**.

When a closed-form solution is unavailable, numerical optimization becomes essential.

## When There Is No Closed-Form Solution

Consider the nonlinear regression model

$$
Y_i
=
\exp(\theta X_i)
+
\varepsilon_i.
$$

Suppose we estimate $\theta$ by minimizing

$$
Q(\theta)
=
\sum_{i=1}^n
\left[
Y_i-\exp(\theta X_i)
\right]^2.
$$

Unlike ordinary linear regression, there is no convenient closed-form least-squares estimator.

```{r}
set.seed(8670)

x <- seq(
  0,
  1,
  length.out = 50
)

y <- exp(1.5 * x) +
  rnorm(
    50,
    sd = 0.15
  )

sse <- function(theta) {
  sum(
    (y - exp(theta * x))^2
  )
}

fit_nonlinear <- optimize(
  sse,
  interval = c(0, 3)
)

fit_nonlinear$minimum
```

```{r}
theta_grid <- seq(
  0,
  3,
  length.out = 400
)

sse_data <- data.frame(
  theta = theta_grid,
  SSE = vapply(
    theta_grid,
    sse,
    numeric(1)
  )
)

ggplot(
  sse_data,
  aes(
    x = theta,
    y = SSE
  )
) +
  geom_line(linewidth = 1) +
  geom_vline(
    xintercept = fit_nonlinear$minimum,
    linetype = "dashed"
  ) +
  labs(
    x = expression(theta),
    y = "Sum of squared errors",
    title = "Optimization for Nonlinear Regression"
  )
```

The important distinction is

$$
\boxed{
\text{Theory tells us what to optimize;}
\qquad
\text{computation tells us how to find it.}
}
$$

## Two Families of Optimization Methods

Computational optimization methods can be broadly divided into two families:

$$
\boxed{
\text{Deterministic Methods}
\qquad\text{and}\qquad
\text{Metaheuristic Methods}.
}
$$

### Deterministic Methods

A deterministic algorithm follows a predictable sequence of updates once the starting value is fixed.

Examples include:

- Newton's method;
- gradient descent;
- quasi-Newton methods such as BFGS;
- deterministic root-finding methods.

Advantages:

- usually efficient for smooth and well-behaved problems;
- mathematically well understood;
- convergence can often be studied theoretically;
- reproducible for a fixed starting value.

Limitations:

- may converge to a local rather than global optimum;
- can be sensitive to the starting value;
- some methods require derivatives;
- poor scaling can substantially slow convergence.

### Metaheuristic Methods

Metaheuristic methods introduce randomness or population-based search to explore the parameter space more broadly.

Examples include:

- simulated annealing;
- genetic algorithms;
- particle swarm optimization.

Advantages:

- can escape local optima;
- usually do not require derivatives;
- useful for multimodal or irregular objectives;
- can be applied to complicated search spaces.

Limitations:

- often computationally expensive;
- results can vary between runs;
- tuning parameters can strongly affect performance;
- global optimality is generally not guaranteed.

The basic trade-off is

$$
\boxed{
\begin{array}{c}
\text{Deterministic}\\
\text{efficient local search}
\end{array}
\qquad\Longleftrightarrow\qquad
\begin{array}{c}
\text{Metaheuristic}\\
\text{broader stochastic search}.
\end{array}
}
$$

## Local and Global Optima

Consider

$$
Q(x)
=
0.05x^2
+
\sin^2(x)
+
0.2\sin(5x).
$$

```{r}
objective <- function(x) {
  0.05 * x^2 +
    sin(x)^2 +
    0.2 * sin(5 * x)
}

x_grid <- seq(
  -7,
  7,
  length.out = 1000
)

objective_data <- data.frame(
  x = x_grid,
  objective = objective(x_grid)
)

ggplot(
  objective_data,
  aes(
    x = x,
    y = objective
  )
) +
  geom_line(linewidth = 1) +
  labs(
    x = "x",
    y = "Objective function",
    title = "Local and Global Optima"
  )
```

The objective contains several local minima.

A deterministic local optimizer may converge to different solutions depending on its starting value.

```{r}
starts <- c(
  -6,
  -2,
  2,
  6
)

solutions <- vapply(
  starts,
  function(x0) {
    optim(
      par = x0,
      fn = objective,
      method = "BFGS"
    )$par
  },
  numeric(1)
)

local_results <- data.frame(
  start = starts,
  solution = solutions,
  objective = objective(solutions)
)

local_results
```

```{r}
ggplot(
  objective_data,
  aes(
    x = x,
    y = objective
  )
) +
  geom_line(linewidth = 1) +
  geom_point(
    data = local_results,
    col = "red",
    aes(
      x = solution,
      y = objective
    ),
    size = 3
  ) +
  labs(
    x = "x",
    y = "Objective function",
    title = "Different Starting Values Can Give Different Solutions"
  )
```

::: {.callout-question title="In-Class Question"}

If the algorithm is deterministic, why can different starting values still produce different answers?

:::

The important distinction is

$$
\boxed{
\text{Local Optimum}
\neq
\text{Global Optimum}.
}
$$

## Deterministic Numerical Methods

We first study several deterministic numerical methods.

### Root Finding

Root finding solves

$$
f(x)=0.
$$

Why is root finding relevant to optimization?

If $Q(x)$ is differentiable and $x^*$ is an interior optimum, then typically

$$
Q'(x^*)=0.
$$

Therefore, optimization can sometimes be converted into

$$
f(x)
=
Q'(x)
=
0.
$$

A root of $Q'(x)$ is only a candidate optimum. It may correspond to a minimum, maximum, or another stationary point.

### A Statistical Example: Normal Quantiles

Suppose we want the 97.5th percentile of a standard normal distribution.

We seek $z$ such that

$$
P(Z\le z)
=
0.975.
$$

Equivalently,

$$
\Phi(z)-0.975=0.
$$

```{r}
f_root <- function(z) {
  pnorm(z) - 0.975
}

z_grid <- seq(
  0,
  4,
  length.out = 400
)

root_data <- data.frame(
  z = z_grid,
  f = f_root(z_grid)
)

ggplot(
  root_data,
  aes(
    x = z,
    y = f
  )
) +
  geom_line(linewidth = 1) +
  geom_hline(
    yintercept = 0,
    linetype = "dashed"
  ) +
  labs(
    x = "z",
    y = expression(Phi(z) - 0.975),
    title = "A Quantile as a Root-Finding Problem"
  )
```

R can solve the root directly:

```{r}
root_R <- uniroot(
  f_root,
  interval = c(0, 4)
)$root

c(
  root_finding = root_R,
  qnorm = qnorm(0.975)
)
```

Thus,

$$
\boxed{
\text{Quantile Calculation}
\Longleftrightarrow
\text{Root Finding}.
}
$$

::: {.callout-question title="In-Class Question"}

Why do `uniroot()` and `qnorm()` give essentially the same answer?

:::

### Bisection Method

The bisection method is similar to a binary search.

Suppose $f$ is continuous and

$$
f(a)f(b)<0.
$$

Then there must be at least one root between $a$ and $b$.

::: {.callout-algorithm title="Bisection Method"}

1. Choose $a$ and $b$ such that $f(a)f(b)<0$.

2. Compute

$$
c=\frac{a+b}{2}.
$$

3. Keep the half containing the sign change.

4. Repeat until the interval is sufficiently small.

:::

The main intuition is

$$
\boxed{
\text{Each iteration cuts the search interval approximately in half}.
}
$$

```{r}
bisect_trace <- function(
    f,
    a,
    b,
    tol = 1e-6,
    maxit = 100
) {

  if (f(a) * f(b) > 0) {
    stop("f(a) and f(b) must have opposite signs.")
  }

  history <- data.frame()

  for (k in 0:maxit) {

    midpoint <- (a + b) / 2

    history <- rbind(
      history,
      data.frame(
        iteration = k,
        a = a,
        b = b,
        midpoint = midpoint,
        f_midpoint = f(midpoint)
      )
    )

    if (
      abs(f(midpoint)) < tol ||
      (b - a) / 2 < tol
    ) {
      break
    }

    if (f(a) * f(midpoint) <= 0) {
      b <- midpoint
    } else {
      a <- midpoint
    }
  }

  history
}

bisect_history <- bisect_trace(
  f_root,
  a = 0,
  b = 4
)

tail(bisect_history)
```

### Visualizing Bisection

```{r}
ggplot(
  bisect_history,
  aes(y = iteration)
) +
  geom_segment(
    aes(
      x = a,
      xend = b,
      yend = iteration
    ),
    linewidth = 2
  ) +
  geom_point(
    aes(x = midpoint),
    size = 2
  ) +
  scale_y_reverse() +
  labs(
    x = "Search interval",
    y = "Iteration",
    title = "Bisection Shrinks the Search Space"
  )
```

Bisection requires continuity and a sign change, but does not require derivatives.

::: {.callout-question title="In-Class Question"}

What information does bisection use about the function?

What information does it ignore?

:::

### Newton-Raphson Method

Newton's method uses local derivative information.

Around the current estimate $x_k$,

$$
f(x)
\approx
f(x_k)
+
f'(x_k)(x-x_k).
$$

Set the approximation equal to zero:

$$
0
=
f(x_k)
+
f'(x_k)(x_{k+1}-x_k).
$$

Therefore,

$$
\boxed{
x_{k+1}
=
x_k
-
\frac{f(x_k)}
{f'(x_k)}
}.
$$

The intuition is:

> Approximate the function locally by a straight line and use the root of that line as the next guess.

For

$$
f(z)
=
\Phi(z)-0.975,
$$

we have

$$
f'(z)
=
\phi(z).
$$

```{r}
f <- function(z) {
  pnorm(z) - 0.975
}

df <- function(z) {
  dnorm(z)
}
```

```{r}
newton_trace <- function(
    f,
    df,
    x0,
    tol = 1e-8,
    maxit = 100
) {

  x <- x0

  history <- data.frame(
    iteration = 0,
    x = x,
    fx = f(x)
  )

  for (k in seq_len(maxit)) {

    dfx <- df(x)

    if (abs(dfx) < 1e-12) {
      stop("Derivative is too close to zero.")
    }

    x_new <- x - f(x) / dfx

    history <- rbind(
      history,
      data.frame(
        iteration = k,
        x = x_new,
        fx = f(x_new)
      )
    )

    if (abs(x_new - x) < tol) {
      break
    }

    x <- x_new
  }

  history
}

newton_history <- newton_trace(
  f,
  df,
  x0 = 1
)

newton_history
```

### Visualizing Newton's Method

```{r}
tangent_data <- data.frame(
  x = head(newton_history$x, -1),
  y = head(newton_history$fx, -1),
  xend = tail(newton_history$x, -1),
  yend = 0
)

z_grid <- seq(
  0.5,
  3,
  length.out = 400
)

newton_function_data <- data.frame(
  z = z_grid,
  f = f(z_grid)
)

ggplot(
  newton_function_data,
  aes(
    x = z,
    y = f
  )
) +
  geom_line(linewidth = 1) +
  geom_hline(
    yintercept = 0,
    linetype = "dashed"
  ) +
  geom_segment(
    data = tangent_data,
    aes(
      x = x,
      y = y,
      xend = xend,
      yend = yend
    ),
    inherit.aes = FALSE,
    linetype = "dashed"
  ) +
  geom_point(
    data = newton_history,
    aes(
      x = x,
      y = fx
    ),
    inherit.aes = FALSE,
    size = 3
  ) +
  labs(
    x = "z",
    y = "f(z)",
    title = "Newton's Method: Follow the Local Tangent"
  )
```

Newton's method can converge very quickly near the root.

However, if

$$
f'(x_k)\approx0,
$$

then the update may become very large and unstable.

::: {.callout-question title="In-Class Question"}

What happens to Newton's update when $f'(x_k)$ is close to zero?

:::

### Secant Method

Newton's method requires the derivative $f'(x)$.

If the derivative is unavailable or expensive, we can approximate it using two previous points:

$$
f'(x_k)
\approx
\frac{
f(x_k)-f(x_{k-1})
}{
x_k-x_{k-1}
}.
$$

Substituting this into Newton's update gives

$$
\boxed{
x_{k+1}
=
x_k
-
f(x_k)
\frac{
x_k-x_{k-1}
}{
f(x_k)-f(x_{k-1})
}
}.
$$

```{r}
secant_trace <- function(
    f,
    x0,
    x1,
    tol = 1e-8,
    maxit = 100
) {

  history <- data.frame(
    iteration = c(0, 1),
    x = c(x0, x1),
    fx = c(
      f(x0),
      f(x1)
    )
  )

  for (k in 2:maxit) {

    f0 <- f(x0)
    f1 <- f(x1)

    denominator <- f1 - f0

    if (abs(denominator) < 1e-12) {
      stop("Secant denominator is too close to zero.")
    }

    x2 <- x1 -
      f1 * (x1 - x0) /
      denominator

    history <- rbind(
      history,
      data.frame(
        iteration = k,
        x = x2,
        fx = f(x2)
      )
    )

    if (abs(x2 - x1) < tol) {
      break
    }

    x0 <- x1
    x1 <- x2
  }

  history
}

secant_history <- secant_trace(
  f,
  x0 = 1,
  x1 = 2.5
)

secant_history
```

The three methods illustrate different computational trade-offs.

| Method | Derivative? | Starting information | Main advantage | Main limitation |
|---|:---:|:---:|---|---|
| Bisection | No | $[a,b]$ | Very reliable | Slower |
| Newton | Yes | $x_0$ | Very fast near root | Can fail |
| Secant | No | $x_0,x_1$ | Fast without derivative | Less reliable |

## Convergence

Suppose an iterative algorithm produces

$$
x_0,x_1,x_2,\ldots
$$

and converges to a solution $x^*$.

Define

$$
e_k
=
|x_k-x^*|.
$$

A common description of convergence is

$$
e_{k+1}
\approx
C e_k^q.
$$

The value $q$ describes how quickly the error decreases.

- **Linear convergence:** $q=1$.
- **Superlinear convergence:** faster than linear.
- **Quadratic convergence:** $q=2$.

Under suitable conditions:

- bisection converges linearly;
- the secant method converges superlinearly;
- Newton's method converges quadratically near the root.

### Comparing Convergence in R

```{r}
true_root <- qnorm(0.975)

bisection_error <- data.frame(
  iteration = bisect_history$iteration,
  error = abs(
    bisect_history$midpoint -
      true_root
  ),
  method = "Bisection"
)

newton_error <- data.frame(
  iteration = newton_history$iteration,
  error = abs(
    newton_history$x -
      true_root
  ),
  method = "Newton"
)

secant_error <- data.frame(
  iteration = secant_history$iteration,
  error = abs(
    secant_history$x -
      true_root
  ),
  method = "Secant"
)

convergence_data <- rbind(
  bisection_error,
  newton_error,
  secant_error
)

convergence_data <- subset(
  convergence_data,
  error > .Machine$double.eps
)
```

```{r}
ggplot(
  convergence_data,
  aes(
    x = iteration,
    y = error,
    linetype = method,
    shape = method
  )
) +
  geom_line() +
  geom_point(size = 2) +
  scale_y_log10() +
  labs(
    x = "Iteration",
    y = "Absolute error",
    title = "Convergence of Root-Finding Methods",
    linetype = "Method",
    shape = "Method"
  )
```

::: {.callout-question title="In-Class Question"}

Which method appears to converge fastest?

Does fewer iterations necessarily mean less computation?

:::

A useful computational perspective is

$$
\boxed{
\text{Total Computational Cost}
\approx
\text{Cost per Iteration}
\times
\text{Number of Iterations}.
}
$$

## From One Dimension to Multiple Dimensions

Most statistical problems involve more than one parameter.

Instead of

$$
x\in\mathbb R,
$$

we now consider

$$
\boldsymbol{\theta}
=
(\theta_1,\ldots,\theta_p)^\top
\in
\mathbb R^p.
$$

The one-dimensional concepts generalize naturally:

$$
Q'(x)
\quad\Longrightarrow\quad
\nabla Q(\boldsymbol{\theta}),
$$

and

$$
Q''(x)
\quad\Longrightarrow\quad
H(\boldsymbol{\theta}).
$$

The **gradient** is

$$
\nabla Q(\boldsymbol{\theta})
=
\begin{pmatrix}
\partial Q/\partial\theta_1\\
\vdots\\
\partial Q/\partial\theta_p
\end{pmatrix},
$$

and the **Hessian** is the matrix of second derivatives,

$$
H(\boldsymbol{\theta})
=
\left[
\frac{
\partial^2 Q
}{
\partial\theta_j\partial\theta_k
}
\right].
$$

### Visualizing a Two-Dimensional Objective

Consider

$$
Q(\theta_1,\theta_2)
=
(\theta_1-2)^2
+
4(\theta_2+1)^2.
$$

```{r}
objective_2d <- function(theta1, theta2) {
  (theta1 - 2)^2 +
    4 * (theta2 + 1)^2
}

theta1_grid <- seq(
  -2,
  6,
  length.out = 150
)

theta2_grid <- seq(
  -4,
  2,
  length.out = 150
)

objective_grid <- expand.grid(
  theta1 = theta1_grid,
  theta2 = theta2_grid
)

objective_grid$value <- with(
  objective_grid,
  objective_2d(theta1, theta2)
)

ggplot(
  objective_grid,
  aes(
    x = theta1,
    y = theta2,
    z = value
  )
) +
  geom_contour(
    bins = 15
  ) +
  geom_point(
    data = data.frame(
      theta1 = 2,
      theta2 = -1
    ),
    aes(
      x = theta1,
      y = theta2
    ),
    inherit.aes = FALSE,
    size = 3
  ) +
  labs(
    x = expression(theta[1]),
    y = expression(theta[2]),
    title = "Contour Plot of a Two-Dimensional Objective"
  )
```

The minimum occurs at

$$
\boldsymbol{\theta}^*
=
\begin{pmatrix}
2\\
-1
\end{pmatrix}.
$$

## Gradient Descent

The gradient points in the direction of greatest increase.

Therefore,

$$
-\nabla Q(\boldsymbol{\theta})
$$

is a natural direction for decreasing the objective.

Gradient descent updates

$$
\boxed{
\boldsymbol{\theta}_{k+1}
=
\boldsymbol{\theta}_k
-
\alpha_k
\nabla Q(\boldsymbol{\theta}_k),
}
$$

where $\alpha_k>0$ is the **step size** or **learning rate**.

For the quadratic example,

$$
\nabla Q(\theta_1,\theta_2)
=
\begin{pmatrix}
2(\theta_1-2)\\
8(\theta_2+1)
\end{pmatrix}.
$$

```{r}
gradient_2d <- function(theta) {
  c(
    2 * (theta[1] - 2),
    8 * (theta[2] + 1)
  )
}

gradient_descent <- function(
    start,
    alpha = 0.1,
    maxit = 30
) {

  theta <- start

  history <- data.frame(
    iteration = 0,
    theta1 = theta[1],
    theta2 = theta[2]
  )

  for (k in seq_len(maxit)) {

    theta <-
      theta -
      alpha * gradient_2d(theta)

    history <- rbind(
      history,
      data.frame(
        iteration = k,
        theta1 = theta[1],
        theta2 = theta[2]
      )
    )
  }

  history
}

gd_path <- gradient_descent(
  start = c(-1, 1),
  alpha = 0.1,
  maxit = 20
)
```

### Visualizing the Optimization Path

```{r}
ggplot(
  objective_grid,
  aes(
    x = theta1,
    y = theta2,
    z = value
  )
) +
  geom_contour(
    bins = 15
  ) +
  geom_path(
    data = gd_path,
    aes(
      x = theta1,
      y = theta2
    ),
    inherit.aes = FALSE
  ) +
  geom_point(
    data = gd_path,
    aes(
      x = theta1,
      y = theta2
    ),
    inherit.aes = FALSE,
    size = 2
  ) +
  labs(
    x = expression(theta[1]),
    y = expression(theta[2]),
    title = "Gradient Descent Path"
  )
```

::: {.callout-question title="In-Class Question"}

What happens if the step size $\alpha$ is too small?

What happens if it is too large?

:::

### Step Size Matters

```{r}
gd_small <- gradient_descent(
  start = c(-1, 1),
  alpha = 0.02,
  maxit = 30
)

gd_medium <- gradient_descent(
  start = c(-1, 1),
  alpha = 0.1,
  maxit = 30
)

gd_large <- gradient_descent(
  start = c(-1, 1),
  alpha = 0.24,
  maxit = 30
)

gd_compare <- rbind(
  transform(
    gd_small,
    step = "alpha = 0.02"
  ),
  transform(
    gd_medium,
    step = "alpha = 0.10"
  ),
  transform(
    gd_large,
    step = "alpha = 0.24"
  )
)

gd_compare$error <-
  sqrt(
    (gd_compare$theta1 - 2)^2 +
      (gd_compare$theta2 + 1)^2
  )

ggplot(
  gd_compare,
  aes(
    x = iteration,
    y = error,
    linetype = step
  )
) +
  geom_line(linewidth = 1) +
  labs(
    x = "Iteration",
    y = "Distance to optimum",
    linetype = "Step size",
    title = "Step Size Changes the Convergence Behaviour"
  )
```

## Multivariate Newton's Method

In one dimension,

$$
x_{k+1}
=
x_k
-
\frac{Q'(x_k)}{Q''(x_k)}.
$$

In multiple dimensions, this becomes

$$
\boxed{
\boldsymbol{\theta}_{k+1}
=
\boldsymbol{\theta}_k
-
H(\boldsymbol{\theta}_k)^{-1}
\nabla Q(\boldsymbol{\theta}_k).
}
$$

However, computationally we usually do **not** calculate the inverse explicitly.

Instead, solve

$$
H(\boldsymbol{\theta}_k)
\boldsymbol{d}_k
=
-\nabla Q(\boldsymbol{\theta}_k),
$$

and update

$$
\boldsymbol{\theta}_{k+1}
=
\boldsymbol{\theta}_k
+
\boldsymbol{d}_k.
$$

::: {.callout-question title="In-Class Question"}

Why do we solve a linear system instead of explicitly computing

$$
H^{-1}?
$$

:::

This connects directly to the earlier discussion of numerical stability and matrix inversion.

## Quasi-Newton Methods

Newton's method requires the Hessian.

For high-dimensional statistical models, computing the Hessian can be expensive.

Quasi-Newton methods approximate curvature information rather than calculating the Hessian directly.

One of the most widely used methods is **BFGS**.

In R:

```r
optim(
  par = start,
  fn = objective,
  method = "BFGS"
)
```

The main computational idea is

$$
\boxed{
\text{Use curvature information}
\quad\text{without repeatedly computing the full Hessian}.
}
$$

## Statistical Example: Logistic Regression

Consider

$$
Y_i
\sim
\operatorname{Bernoulli}(p_i),
$$

with

$$
\log
\left(
\frac{p_i}{1-p_i}
\right)
=
\beta_0+\beta_1x_i.
$$

The parameter is now a vector

$$
\boldsymbol{\beta}
=
(\beta_0,\beta_1)^\top.
$$

There is no closed-form MLE, so logistic regression requires numerical optimization.

```{r}
set.seed(8670)

n <- 200

x_logit <- rnorm(n)

beta_true <- c(
  -0.5,
  1.5
)

eta <-
  beta_true[1] +
  beta_true[2] * x_logit

p <- plogis(eta)

y_logit <- rbinom(
  n,
  size = 1,
  prob = p
)
```

Define the negative log-likelihood:

```{r}
logistic_nll <- function(beta) {

  eta <-
    beta[1] +
    beta[2] * x_logit

  p <- plogis(eta)

  -sum(
    dbinom(
      y_logit,
      size = 1,
      prob = p,
      log = TRUE
    )
  )
}
```

Estimate the coefficients using BFGS:

```{r}
fit_optim <- optim(
  par = c(0, 0),
  fn = logistic_nll,
  method = "BFGS"
)

fit_optim$par
```

Compare with `glm()`:

```{r}
fit_glm <- glm(
  y_logit ~ x_logit,
  family = binomial()
)

coef(fit_glm)

fit_optim$par
```

::: {.callout-question title="In-Class Question"}

Why is comparing `optim()` with `glm()` useful?

:::

This illustrates the complete computational-statistics workflow:

$$
\boxed{
\text{Statistical Model}
\rightarrow
\text{Likelihood}
\rightarrow
\text{Optimization}
\rightarrow
\text{Algorithm}
\rightarrow
\text{Validation}.
}
$$

## Scaling and Conditioning

Optimization can behave very differently when parameters operate on different scales.

Consider

$$
Q(\theta_1,\theta_2)
=
(\theta_1-1)^2
+
100(\theta_2-1)^2.
$$

```{r}
scaled_objective <- function(theta1, theta2) {
  (theta1 - 1)^2 +
    100 * (theta2 - 1)^2
}

scale_grid <- expand.grid(
  theta1 = seq(
    -2,
    4,
    length.out = 150
  ),
  theta2 = seq(
    -1,
    3,
    length.out = 150
  )
)

scale_grid$value <- with(
  scale_grid,
  scaled_objective(
    theta1,
    theta2
  )
)

ggplot(
  scale_grid,
  aes(
    x = theta1,
    y = theta2,
    z = value
  )
) +
  geom_contour(
    bins = 15
  ) +
  labs(
    x = expression(theta[1]),
    y = expression(theta[2]),
    title = "Poor Scaling Produces an Elongated Objective"
  )
```

A highly elongated objective can cause gradient-based algorithms to zig-zag and converge slowly.

This illustrates another important principle:

$$
\boxed{
\text{Same mathematical problem}
+
\text{different scaling}
\Longrightarrow
\text{different computational behaviour}.
}
$$

## Metaheuristic Methods

Deterministic optimization methods can work extremely well for smooth and well-behaved objectives.

However, they may struggle when the objective:

- contains many local optima;
- is not differentiable;
- is discontinuous;
- has a complicated search space.

Metaheuristic methods use broader stochastic exploration.

The key idea is

$$
\boxed{
\text{Explore broadly}
\quad\longrightarrow\quad
\text{search more locally later}.
}
$$

### Why Accept a Worse Solution?

A local deterministic method usually attempts to improve the objective at every step.

A stochastic method may deliberately accept a worse solution.

::: {.callout-question title="In-Class Question"}

Why would an optimization algorithm ever accept a worse solution?

:::

Temporarily moving to a worse location can allow the algorithm to escape a local optimum and later reach a better region.

## Simulated Annealing

Simulated annealing is a stochastic optimization method inspired by the physical process of heating and slowly cooling a material.

Suppose the current solution is $x_k$ and a candidate solution $x'$ is generated.

For a minimization problem, if

$$
Q(x')<Q(x_k),
$$

the new solution is accepted.

If

$$
Q(x')>Q(x_k),
$$

the worse solution may still be accepted with probability

$$
\boxed{
P(\text{accept})
=
\exp
\left(
-\frac{\Delta Q}{T}
\right),
}
$$

where

$$
\Delta Q
=
Q(x')-Q(x_k)>0.
$$

The parameter $T$ is called the **temperature**.

When

$$
T
\quad\text{is large},
$$

the algorithm explores more broadly.

When

$$
T
\quad\text{is small},
$$

the algorithm becomes more selective.

### Visualizing the Acceptance Probability

```{r}
delta_grid <- seq(
  0,
  5,
  length.out = 300
)

temperature <- c(
  0.1,
  0.5,
  2
)

sa_probability <- expand.grid(
  delta = delta_grid,
  temperature = temperature
)

sa_probability$probability <-
  exp(
    -sa_probability$delta /
      sa_probability$temperature
  )

sa_probability$temperature <-
  factor(
    sa_probability$temperature
  )

ggplot(
  sa_probability,
  aes(
    x = delta,
    y = probability,
    linetype = temperature
  )
) +
  geom_line(linewidth = 1) +
  labs(
    x = expression(Delta * Q),
    y = "Probability of accepting a worse solution",
    linetype = "Temperature",
    title = "Temperature Controls Exploration"
  )
```

::: {.callout-question title="In-Class Question"}

At high temperature, why is the algorithm more willing to accept a poor move?

What happens as the temperature approaches zero?

:::

### Simulated Annealing in R

R provides simulated annealing through

```r
optim(
  ...,
  method = "SANN"
)
```

Using the earlier multimodal objective:

```{r}
set.seed(8670)

sa_fit <- optim(
  par = 5,
  fn = objective,
  method = "SANN",
  control = list(
    maxit = 3000
  )
)

sa_fit$par
sa_fit$value
```

Because simulated annealing is stochastic, different seeds may produce different results.

```{r}
seeds <- 1:20

sa_solution <- vapply(
  seeds,
  function(seed) {

    set.seed(seed)

    optim(
      par = 5,
      fn = objective,
      method = "SANN",
      control = list(
        maxit = 3000
      )
    )$par

  },
  numeric(1)
)

sa_data <- data.frame(
  seed = seeds,
  solution = sa_solution,
  objective = objective(sa_solution)
)

ggplot(
  sa_data,
  aes(
    x = factor(seed),
    y = objective
  )
) +
  geom_point(size = 3) +
  labs(
    x = "Random seed",
    y = "Final objective value",
    title = "Variation Across Simulated Annealing Runs"
  )
```

### Advantages of Simulated Annealing

- Can escape local optima.
- Does not require derivatives.
- Easy to adapt to many objective functions.
- Useful for multimodal problems.

### Limitations of Simulated Annealing

- Can require many function evaluations.
- Results vary across runs.
- Performance depends on the cooling schedule.
- Does not guarantee the global optimum.

## Genetic Algorithm

A genetic algorithm uses a **population of candidate solutions** rather than one current solution.

The basic cycle is

$$
\boxed{
\text{Selection}
\rightarrow
\text{Crossover}
\rightarrow
\text{Mutation}
\rightarrow
\text{New Generation}.
}
$$

### Selection

Better candidate solutions are more likely to remain in the population.

### Crossover

Information from two candidate solutions is combined to create new candidates.

### Mutation

Random changes are introduced to preserve diversity and explore new regions.

The key difference from simulated annealing is:

$$
\boxed{
\text{SA: one evolving solution}
\qquad
\text{GA: a population of solutions}.
}
$$

### Advantages

- Explores several regions simultaneously.
- Does not require gradients.
- Flexible for continuous and discrete problems.

### Limitations

- Can require many objective evaluations.
- Requires tuning of population size and mutation rate.
- Can converge slowly.
- Does not guarantee a global optimum.

## Particle Swarm Optimization

Particle swarm optimization also maintains a population of candidate solutions called **particles**.

Each particle uses information about:

- its current location;
- its previous velocity;
- its own best location;
- the best location found by the group.

A simplified velocity update is

$$
v_{i,k+1}
=
w v_{i,k}
+
c_1r_1(p_i-x_{i,k})
+
c_2r_2(g-x_{i,k}),
$$

followed by

$$
x_{i,k+1}
=
x_{i,k}
+
v_{i,k+1}.
$$

The intuition is

$$
\boxed{
\text{Momentum}
+
\text{Individual Experience}
+
\text{Group Experience}
+
\text{Randomness}.
}
$$

### Advantages

- Does not require derivatives.
- Explores multiple regions simultaneously.
- Relatively intuitive and flexible.

### Limitations

- May require many function evaluations.
- Sensitive to tuning parameters.
- Can converge prematurely.
- Global optimality is not guaranteed.

## Deterministic versus Metaheuristic Methods

The two families search very differently.

| Method | Type | Derivatives | Search | Main Advantage | Main Limitation |
|---|---|:---:|---|---|---|
| Bisection | Deterministic | No | Root finding | Very reliable | Slow |
| Newton | Deterministic | Yes | Local | Very fast near solution | Sensitive to derivative and start |
| Secant | Deterministic | No | Local | Fast without derivative | Less reliable |
| Gradient Descent | Deterministic | Gradient | Local | Simple and scalable | Step size matters |
| BFGS | Deterministic | Usually gradient | Local | Efficient for smooth problems | Can reach local optimum |
| Simulated Annealing | Metaheuristic | No | Global exploration | Can escape local optima | Slow and stochastic |
| Genetic Algorithm | Metaheuristic | No | Population | Flexible search | Computationally expensive |
| Particle Swarm | Metaheuristic | No | Population | Cooperative exploration | Requires tuning |

The important question is not

> Which algorithm is the best?

Instead, ask

> Which algorithm is appropriate for the structure of the statistical problem?

For a smooth and well-behaved objective,

$$
\text{deterministic methods}
$$

are usually preferable.

For a highly multimodal, irregular, or derivative-free objective,

$$
\text{metaheuristic methods}
$$

may become useful.

::: {.callout-question title="In-Class Question"}

Suppose the objective is smooth, convex, and easy to differentiate.

Would you choose BFGS or simulated annealing?

What changes if the objective has many local minima and no useful derivatives?

:::

## Optimization Transfer

Sometimes the original objective is difficult to optimize directly.

Instead, we construct an easier **surrogate function** and repeatedly optimize the surrogate.

The general idea is

$$
\boxed{
\text{Difficult Objective}
\rightarrow
\text{Easier Surrogate}
\rightarrow
\text{Update}
\rightarrow
\text{Repeat}.
}
$$

This principle is often called **optimization transfer**.

The surrogate should be easier to optimize while still guiding the algorithm toward improvement of the original objective.

We will encounter this idea again in later statistical algorithms.

## Computational Perspective

The goal is not simply to memorize optimization algorithms.

For a statistical problem, ask:

1. What is the statistical quantity we want?
2. Can it be written as an optimization or root-finding problem?
3. Is there a closed-form solution?
4. Is the parameter scalar or multivariate?
5. Is the objective smooth?
6. Are derivatives available?
7. Are there many local optima?
8. How sensitive is the method to the starting value?
9. Is the computation numerically stable?
10. How can we validate the result?

This brings us back to the central idea of statistical computing:

$$
\boxed{
\text{Theory}
\Longleftrightarrow
\text{Computation}
\Longleftrightarrow
\text{Practice}.
}
$$

## Questions to Think at Home

1. Why can a mathematically correct optimization formula still lead to a poor computational implementation?

2. Why is a closed-form solution useful even when numerical optimization will eventually be used?

3. Why is bisection more reliable than Newton's method, and what is the computational cost of that reliability?

4. What additional information does Newton's method use compared with bisection?

5. Why can different starting values lead to different solutions for the same objective function?

6. What role does the step size play in gradient descent?

7. Why might poor scaling make optimization difficult?

8. Why is solving

$$
H\boldsymbol{d}
=
-\nabla Q
$$

usually preferable to explicitly computing $H^{-1}$?

9. Why can BFGS be useful when the Hessian is expensive to compute?

10. Why might a metaheuristic algorithm deliberately accept a worse solution?

11. When would you prefer simulated annealing over a deterministic local optimizer?

12. What is the main difference between simulated annealing, genetic algorithms, and particle swarm optimization?

13. Why does fewer iterations not necessarily imply lower computational cost?

14. How are optimization, likelihood estimation, and statistical inference connected?

## Takeaways

- Many statistical estimators are solutions to optimization problems.
- Closed-form solutions are convenient but often unavailable.
- Numerical optimization connects statistical theory with practical computation.
- Root finding is closely related to optimization.
- Bisection is reliable but relatively slow.
- Newton's method is fast near a solution but uses derivative information.
- The secant method approximates derivative information numerically.
- Multivariate optimization replaces derivatives with gradients and Hessians.
- Gradient descent uses local first-order information.
- Newton's method uses curvature information through the Hessian.
- BFGS approximates curvature without repeatedly computing the full Hessian.
- Scaling can strongly affect computational performance.
- Deterministic methods are usually efficient for smooth, well-behaved objectives.
- Metaheuristic methods trade speed and predictability for broader exploration.
- Simulated annealing can escape local optima by occasionally accepting worse solutions.
- Genetic algorithms and particle swarm optimization use populations of candidate solutions.
- There is no universally best optimization algorithm.
- A good computational method balances accuracy, stability, speed, and exploration.

---

Examples are adapted from the following sources:

- Peng, R.D. [Advanced Statistical Computing](https://bookdown.org/rdpeng/advstatcomp/).