# Generating Random Variables {#sec-generating-rv}

<!-- Adapted and expanded from 03-generating-rv.qmd in the master branch of
     chikuang/Teaching-Stat8670 for the 2026F course. -->


```{r}
#| label: rv-setup
#| include: false
#| code-fold: true
library(ggplot2)
theme_set(theme_minimal(base_size = 14))
knitr::opts_chunk$set(
  fig.width = 8, fig.height = 4.8, fig.align = "center",
  dpi = 144, collapse = TRUE, comment = "#>"
)
```

In the previous chapter, we used optimization to find a parameter value that fits observed data. In this chapter, we move in the other direction: **given a probability model, how can we generate data that behave according to that model?**

These two directions work together. We can fit a model to observations, generate new data from the fitted model, and ask whether the simulated data reproduce important features of the observations.

The central idea is

$$
\text{Probability Model}
\;\Longrightarrow\;
\text{Random Data}
\;\Longrightarrow\;
\text{Statistical Procedure}
\;\Longrightarrow\;
\text{Operating Characteristics}.
$$

Here, *operating characteristics* include bias, variability, confidence interval coverage, and the probability of rejecting a hypothesis.

## Why Generate Random Variables?

Suppose we propose an estimator $\hat\theta$ for an unknown parameter $\theta$. For one observed dataset, we cannot see the estimator's entire sampling distribution. Under a specified model, however, we can repeatedly generate datasets with a known value of $\theta$ and recompute $\hat\theta$.

This lets us investigate questions that may be difficult to answer analytically.

| Statistical question | What must be generated? | Why the generator matters |
|---|---|---|
| How variable is an estimator? | Repeated datasets from a specified model | The simulated sampling distribution depends on that model. |
| Does a 95% confidence interval cover 95% of the time? | Data with a known true parameter | We can count intervals that contain the truth. |
| How much data does an experiment need? | Data under meaningful alternatives | The fraction of rejections estimates power. |
| How uncertain is a fitted prediction? | New observations, possibly with uncertain parameters | Both observation noise and parameter uncertainty can matter. |
| What does a bootstrap approximate? | Samples from an empirical or fitted distribution | Resampling defines the distribution used to approximate uncertainty. |
| How does dependence affect a joint event? | Entire random vectors | Correct marginal distributions alone do not specify joint probabilities. |

**Statistical intuition.** Simulation gives us a laboratory in which the truth is known. We control the data-generating mechanism and observe how a method behaves. The conclusions apply to the mechanisms we examine; simulating one convenient model does not establish that a procedure works under every model.

::: {.callout-question title="In-Class Question"}

An estimator performs well when data are normal. What additional simulations would you want before using it with skewed data, outliers, or dependent observations?

:::

## Uniform Random Numbers as Building Blocks

### Pseudorandomness and Reproducibility

A pseudorandom number generator produces a deterministic sequence from an initial state. A well-designed generator makes that sequence useful as an approximation to independent random draws. In the mathematical arguments below, we work with ideal independent uniforms; software approximates them with finite precision.

In R, use `set.seed()` to reproduce an example. The same seed, generator settings, software environment, and sequence of calls reproduce the same simulation. The seed identifies a computation; it does not certify the probability model.

```{r}
#| label: rv-seed
#| code-fold: true
set.seed(8670)
u_first <- runif(5)

set.seed(8670)
u_repeated <- runif(5)

rbind(first = u_first, repeated = u_repeated)
identical(u_first, u_repeated)
```

Set the seed **before a simulation experiment**, then let the generator advance. Resetting the same seed inside every replication creates repeated copies of the same random input.

For a reproducible analysis, also record `RNGkind()` and `sessionInfo()`. For parallel work, use a framework that manages separate random-number streams, such as the streams supported by R's `parallel` package. See the [R documentation on random-number generation](https://stat.ethz.ch/R-manual/R-devel/library/base/html/Random.html).

### Sampling from a Finite Population

Sampling with replacement allows an observation to appear more than once. Sampling without replacement makes later draws depend on which observations have already been selected.

```{r}
#| label: rv-sampling
#| code-fold: true
set.seed(8671)
observed_waits <- c(2, 3, 3, 5, 7, 12)  # illustrative waiting times

sample(observed_waits, size = 6, replace = TRUE)
sample(observed_waits, size = 6, replace = FALSE)
```

The empirical distribution places probability $1/n$ on each of the $n$ observed records, counting repeated values separately. Drawing $n$ times independently from this distribution is the basic **nonparametric bootstrap** resampling step. Sampling all $n$ records without replacement only permutes them, so the sample mean cannot change.

This is our first example of an important principle: **the sampling rule is part of the statistical model.**

### R's Distribution Functions

R uses four related prefixes. For example, `dnorm()` evaluates a normal density, `pnorm()` evaluates its CDF, `qnorm()` evaluates its quantile function, and `rnorm()` generates draws.

| Distribution | Density or PMF | CDF | Quantile | Generator | Parameters to check |
|---|---|---|---|---|---|
| Uniform | `dunif` | `punif` | `qunif` | `runif` | `min`, `max` |
| Normal | `dnorm` | `pnorm` | `qnorm` | `rnorm` | `mean`, **`sd`** |
| Exponential | `dexp` | `pexp` | `qexp` | `rexp` | **`rate`** |
| Gamma | `dgamma` | `pgamma` | `qgamma` | `rgamma` | `shape`, `rate` or `scale` |
| Beta | `dbeta` | `pbeta` | `qbeta` | `rbeta` | `shape1`, `shape2` |
| Binomial | `dbinom` | `pbinom` | `qbinom` | `rbinom` | `size`, `prob` |
| Geometric | `dgeom` | `pgeom` | `qgeom` | `rgeom` | `prob`; counts failures |
| Poisson | `dpois` | `ppois` | `qpois` | `rpois` | `lambda` |
| Negative binomial | `dnbinom` | `pnbinom` | `qnbinom` | `rnbinom` | `size`, `prob` or `mu` |
| Chi-square | `dchisq` | `pchisq` | `qchisq` | `rchisq` | `df` |
| Student's $t$ | `dt` | `pt` | `qt` | `rt` | `df` |
| $F$ | `df` | `pf` | `qf` | `rf` | `df1`, `df2` |
| Lognormal | `dlnorm` | `plnorm` | `qlnorm` | `rlnorm` | `meanlog`, `sdlog` |

See [R's distribution reference](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/Distributions.html) for the available families.

::: {.callout-note title="Parameters Have Statistical Meaning"}

`rnorm(n, sd = 4)` has variance $16$. An exponential rate of $4$ gives mean $1/4$. In the gamma family, `scale = 1 / rate`. A lognormal `meanlog` is the mean of $\log X$, not the mean of $X$.

A generator can run without errors while representing the wrong model if its parameterization is misunderstood.

:::

## Inverse Transformation Method

### Turning Probability Levels into Observations

Imagine dividing the interval $(0,1)$ into probability levels. A uniform draw selects a level without favoring one part of that interval. The quantile function converts that probability level into a value on the scale of the target distribution.

::: {.callout-theorem title="Inverse Transformation"}

For any CDF $F$, define its generalized inverse by

$$
F^{-1}(u)=\inf\{x\in\mathbb{R}:F(x)\geq u\},
\qquad 0<u<1.
$$

If $U\sim\operatorname{Unif}(0,1)$, then $X=F^{-1}(U)$ has CDF $F$.

:::

The inequality **$F(x)\geq u$** matters. For a discrete distribution, the CDF jumps and may never equal a particular value of $u$.

To understand why the method works, observe that

$$
\begin{aligned}
P\{F^{-1}(U)\leq x\}
&=P\{U\leq F(x)\}\\
&=F(x).
\end{aligned}
$$

Thus the transformation produces exactly the required probabilities in ideal arithmetic.

The reverse statement is the **probability integral transform**: if $X$ has a continuous CDF $F$, then $F(X)\sim\operatorname{Unif}(0,1)$. This gives a connection to model checking. Under a correctly specified continuous predictive distribution, transformed observations should have a uniform marginal distribution. Dependence must be checked separately, and estimating $F$ from the same observations changes the reference distribution of formal tests.

::: {.callout-algorithm title="Inverse Transformation"}

1. Obtain the quantile function $F^{-1}$ analytically or numerically.
2. Generate independent $U_1,\ldots,U_n\sim\operatorname{Unif}(0,1)$.
3. Return $X_i=F^{-1}(U_i)$ for each $i$.

:::

### Continuous Example: A Density That Increases

::: {.callout-example title="An Increasing Density on the Unit Interval"}

Consider

$$
f(x)=3x^2,\qquad 0<x<1.
$$

Within the support, $F(x)=x^3$, so $F^{-1}(u)=u^{1/3}$. Therefore,

$$
X=U^{1/3}.
$$

Why do the observations concentrate near 1? Equal-width intervals near 1 contain more probability because the CDF rises faster there. The inverse maps more uniform probability levels into those intervals.

```{r}
#| label: fig-rv-inverse
#| code-fold: true
#| fig-cap: "Left: uniform probability levels. Right: mapping them through the quantile function produces values with the increasing target density."
#| fig-width: 9
set.seed(8672)
n <- 5000
u <- runif(n)
x_power <- u^(1 / 3)

plot_data <- rbind(
  data.frame(value = u, distribution = "1. Uniform input"),
  data.frame(value = x_power, distribution = "2. Transformed output")
)
grid <- seq(0, 1, length.out = 301)
curve_data <- rbind(
  data.frame(value = grid, density = 1,
             distribution = "1. Uniform input"),
  data.frame(value = grid, density = 3 * grid^2,
             distribution = "2. Transformed output")
)

ggplot(plot_data, aes(value)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30,
                 boundary = 0, fill = "lightblue", color = "white") +
  geom_line(data = curve_data, aes(y = density),
            color = "firebrick", linewidth = 1) +
  facet_wrap(~ distribution) +
  labs(x = "Value", y = "Density")
```

Here $\mathbb{E}(X)=3/4$ and $\operatorname{Var}(X)=3/80$. A simulation provides an independent numerical check.

```{r}
#| label: rv-power-moments
#| code-fold: true
knitr::kable(
  data.frame(
    quantity = c("Mean", "Variance"),
    simulated = c(mean(x_power), var(x_power)),
    theoretical = c(3 / 4, 3 / 80)
  ),
  digits = 4
)
```

Agreement should be approximate. Random samples are not required to reproduce population moments exactly.

:::

### Exponential Waiting Times

For $T\sim\operatorname{Exp}(\lambda)$ with **rate** $\lambda>0$,

$$
F(t)=1-e^{-\lambda t},
\qquad
F^{-1}(u)=-\frac{\log(1-u)}{\lambda}.
$$

Hence we can generate waiting times using

$$
T=-\frac{\log(1-U)}{\lambda}.
$$

Since $U$ and $1-U$ have the same distribution, $-\log(U)/\lambda$ also works. These two formulas have the same marginal distribution; applying both to the **same** $U$ produces dependent outputs.

```{r}
#| label: rv-exponential
#| code-fold: true
set.seed(8673)
lambda <- 0.7
u <- runif(10000)
waiting_time <- -log1p(-u) / lambda

c(
  simulated_mean = mean(waiting_time),
  theoretical_mean = 1 / lambda,
  simulated_prob_above_3 = mean(waiting_time > 3),
  theoretical_prob_above_3 = exp(-lambda * 3)
)
```

`log1p(-u)` computes $\log(1-u)$ accurately when $u$ is close to zero. This connects random-number generation to the numerical stability issues discussed earlier.

**Why this matters.** Exponential waiting times model the time between events in a homogeneous Poisson process. The constant rate is a substantive assumption: it implies a memoryless waiting time. Equipment aging or time-varying arrival rates may require a different distribution.

More generally, if a nonnegative event time has continuous cumulative hazard $H(t)$ and survival function $S(t)=e^{-H(t)}$ tending to zero, then

$$
T=H^{-1}\{-\log(U)\}.
$$

This turns survival-model simulation into an inversion problem. For example, the Weibull cumulative hazard $H(t)=(t/b)^a$ gives $T=b\{-\log(U)\}^{1/a}$.

::: {.callout-question title="In-Class Question"}

If the exponential rate doubles, what happens to the average waiting time? Why would multiplying the simulated waiting times by two represent the opposite change?

:::

### Discrete Distributions

For ordered support values $x_1<x_2<\cdots$, the quantile rule selects the first value whose cumulative probability reaches $u$:

$$
X=x_j
\quad\text{when}\quad
F(x_{j-1})<U\leq F(x_j).
$$

Each outcome receives a subinterval of $(0,1)$ whose length equals its probability.

::: {.callout-example title="Bernoulli and Categorical Outcomes"}

For $X\sim\operatorname{Bernoulli}(p)$, the generalized inverse gives

$$
X=\mathbf{1}\{U>1-p\}.
$$

The threshold divides the unit interval into pieces of lengths $1-p$ and $p$.

```{r}
#| label: rv-discrete
#| code-fold: true
set.seed(8674)
p <- 0.4
bernoulli <- as.integer(runif(10000) > 1 - p)

labels <- c("Low", "Medium", "High")
probabilities <- c(0.2, 0.5, 0.3)
cutoffs <- cumsum(probabilities)
u <- runif(10000)

# First cutoff that reaches u: this also handles u at a cutoff.
index <- vapply(u, function(ui) which(ui <= cutoffs)[1], integer(1))
category <- factor(labels[index], levels = labels)

mean(bernoulli)
prop.table(table(category))

# Built-in alternative for the same categorical distribution.
head(sample(labels, size = 10, replace = TRUE, prob = probabilities))
```

:::

For the geometric distribution, use R's convention: $X$ counts **failures before the first success**. Then

$$
P(X=x)=p(1-p)^x,\qquad x=0,1,2,\ldots,
$$

and $F(x)=1-(1-p)^{x+1}$. Solving $F(x-1)<u\leq F(x)$ gives

$$
F^{-1}(u)=
\left\lceil\frac{\log(1-u)}{\log(1-p)}\right\rceil-1.
$$

```{r}
#| label: rv-geometric
#| code-fold: true
set.seed(8675)
p <- 0.25
u <- runif(10000)
failures <- ceiling(log1p(-u) / log1p(-p)) - 1

c(
  simulated_mean = mean(failures),
  theoretical_mean = (1 - p) / p,
  largest_difference_from_qgeom = max(abs(failures - qgeom(u, prob = p)))
)
```

The number of **trials including the success** is $X+1$, with mean $1/p$. Mixing up these conventions creates an off-by-one modeling error. See [R's geometric distribution documentation](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/Geometric.html).

For a Poisson distribution, cumulative probabilities can be constructed from

$$
p_0=e^{-\lambda},\qquad
p_{k+1}=\frac{\lambda}{k+1}p_k,
\qquad
F(k+1)=F(k)+p_{k+1}.
$$

Search for the first $k$ satisfying $F(k)\geq u$. This explains an inversion algorithm, but the recurrence can underflow for large $\lambda$. In practice, use `qpois(runif(n), lambda)` or the specialized generator `rpois(n, lambda)`. An `r` function need not implement the simple algorithm used in our derivation.

::: {.callout-note title="The Probability Integral Transform Has a Continuity Condition"}

For discrete $X$, $F(X)$ takes only a discrete set of values and is not uniform. One randomized version is

$$
U=F(X^-)+V\{F(X)-F(X^-)\},
$$

where $V$ is an independent uniform and $F(X^-)$ is the CDF just below $X$. This spreads each probability mass over its corresponding interval.

:::

### Numerical Inversion: A Link to Root Finding

An inverse may be inconvenient algebraically even when the CDF is easy to evaluate. For each $u$, solve

$$
F(x)-u=0.
$$

For example, consider

$$
F(x)=\frac{x+x^3}{2},\quad 0\leq x\leq 1,
\qquad
f(x)=\frac{1+3x^2}{2}.
$$

The CDF is strictly increasing, so each $u\in(0,1)$ has exactly one root in $(0,1)$. We can use the bracketing methods from the optimization chapter.

```{r}
#| label: rv-numerical-inversion
#| code-fold: true
custom_cdf <- function(x) (x + x^3) / 2  # used only on [0, 1]

custom_quantile <- function(u) {
  vapply(u, function(ui) {
    uniroot(
      function(x) custom_cdf(x) - ui,
      interval = c(0, 1), tol = 1e-10
    )$root
  }, numeric(1))
}

set.seed(8676)
u <- runif(2000)
x_numeric <- custom_quantile(u)

c(
  largest_cdf_residual = max(abs(custom_cdf(x_numeric) - u)),
  simulated_mean = mean(x_numeric),
  theoretical_mean = 5 / 8
)
```

There are two distinct errors: the root solver introduces **numerical error**, while finite simulation introduces **Monte Carlo error**. Tightening `tol` reduces the former; increasing the number of independent draws reduces the latter. Neither repairs a wrong CDF.

### Conditional Simulation: Truncation

Suppose only values in $(a,b)$ are eligible for a study. If $F$ is continuous and $F(b)>F(a)$, then the conditional CDF within this interval is

$$
P(X\leq x\mid a<X<b)
=\frac{F(x)-F(a)}{F(b)-F(a)}.
$$

Consequently,

$$
X=F^{-1}\left[F(a)+U\{F(b)-F(a)\}\right]
$$

generates from the truncated distribution.

```{r}
#| label: rv-truncation
#| code-fold: true
set.seed(8677)
a <- 1
b <- 2
u <- runif(5000)
truncated_normal <- qnorm(pnorm(a) + u * (pnorm(b) - pnorm(a)))

c(
  simulated_mean = mean(truncated_normal),
  theoretical_mean = (dnorm(a) - dnorm(b)) / (pnorm(b) - pnorm(a)),
  fraction_in_interval = mean(truncated_normal > a & truncated_normal < b)
)
```

**Practical interpretation.** Truncation changes which individuals enter the dataset. Censoring instead records incomplete information about some individuals who remain in the dataset, such as knowing that a failure time exceeds the study duration. These require different simulation rules. The direct CDF subtraction above is suitable for this moderate interval; extreme tails require numerically stable tail probabilities or a specialized truncated-distribution generator.

## Acceptance-Rejection Method

### Reshaping an Easy Distribution

Inverse transformation requires a usable quantile function. Acceptance-rejection requires a different resource: an easy **proposal distribution** that can cover the target.

Let $f$ be the normalized target density and $g$ a proposal density. We need a finite constant $M$ such that

$$
f(y)\leq M g(y)
$$

everywhere on the target support. In particular, $g(y)$ must be positive wherever $f(y)>0$.

::: {.callout-algorithm title="Acceptance-Rejection Sampling"}

Repeat until the desired number of observations has been accepted:

1. Generate $Y\sim g$.
2. Independently generate $U\sim\operatorname{Unif}(0,1)$.
3. Accept $Y$ if $U\leq f(Y)/\{M g(Y)\}$; otherwise discard it.

:::

**Statistical intuition.** The proposal visits some regions too frequently. The acceptance rule thins those visits by a location-specific amount. After thinning, the relative frequencies match the target.

### Why Does It Work?

For continuous proposals,

$$
P(\text{accept})
=\int g(y)\frac{f(y)}{M g(y)}\,dy
=\frac{1}{M}.
$$

The density of an accepted proposal is therefore

$$
f_{Y\mid\text{accept}}(y)
=\frac{g(y)P(\text{accept}\mid Y=y)}
        {P(\text{accept})}
=f(y).
$$

For discrete distributions, replace the integral by a sum. Independent proposal/acceptance pairs produce independent accepted draws. This differs from a Markov chain method, in which successive draws generally depend on the current state.

The number of proposals through one acceptance is geometric on $\{1,2,\ldots\}$ with mean $M$. To obtain $n$ accepted observations requires **$nM$ proposals on average**, not exactly $nM$ proposals.

### A Beta Example: A Good Bound Saves Work

For the $\operatorname{Beta}(2,2)$ target,

$$
f(y)=6y(1-y),\qquad 0<y<1.
$$

With a uniform proposal, $g(y)=1$. Since the maximum occurs at $y=1/2$,

$$
M_{\min}=\max_{0<y<1}6y(1-y)=\frac{3}{2}.
$$

The resulting acceptance probability is $2/3$. Choosing $M=6$ is also valid, but accepts only $1/6$ of proposals on average. A bound can be correct and unnecessarily expensive.

```{r}
#| label: fig-rv-rejection
#| code-fold: true
#| fig-cap: "Uniform proposals form points under the envelope of height 1.5. Retaining only points below the target density produces Beta(2,2) draws along the horizontal axis."
set.seed(8678)
proposal <- runif(1500)
height <- 1.5 * runif(1500)
accepted <- height <= dbeta(proposal, 2, 2)
proposal_data <- data.frame(proposal, height, accepted)

ggplot(proposal_data, aes(proposal, height, color = accepted)) +
  geom_point(alpha = 0.6, size = 1.2) +
  stat_function(fun = function(x) dbeta(x, 2, 2),
                color = "black", linewidth = 1) +
  geom_hline(yintercept = 1.5, linetype = "dashed") +
  scale_color_manual(values = c("grey70", "steelblue"),
                     labels = c("Rejected", "Accepted")) +
  labs(x = "Proposal y", y = "Height", color = NULL)
```

The following implementation stops as soon as it has exactly $n$ accepted observations. Counting all proposals makes its efficiency visible.

```{r}
#| label: rv-rejection-generator
#| code-fold: true
rbeta22_rejection <- function(n, M = 1.5) {
  stopifnot(
    length(n) == 1L, is.finite(n), n >= 0, n == floor(n),
    length(M) == 1L, is.finite(M), M >= 1.5
  )
  draws <- numeric(n)
  n_accepted <- 0L
  proposals <- 0L

  while (n_accepted < n) {
    y <- runif(1)
    u <- runif(1)
    proposals <- proposals + 1L
    if (u <= dbeta(y, 2, 2) / M) {
      n_accepted <- n_accepted + 1L
      draws[n_accepted] <- y
    }
  }
  list(draws = draws, proposals = proposals)
}

set.seed(8679)
beta_tight <- rbeta22_rejection(2000, M = 1.5)
beta_loose <- rbeta22_rejection(2000, M = 6)

knitr::kable(data.frame(
  M = c(1.5, 6),
  proposals = c(beta_tight$proposals, beta_loose$proposals),
  expected_proposals = 2000 * c(1.5, 6),
  observed_acceptance = 2000 / c(beta_tight$proposals, beta_loose$proposals),
  theoretical_acceptance = 1 / c(1.5, 6)
), digits = 3, col.names = c(
  "M", "Proposals", "Expected proposals", "Observed rate", "Target rate"
))
```

Each proposal here uses two uniforms, one for $Y$ and one for the decision. Thus $n=2000$ with $M=1.5$ uses $3000$ proposals and $6000$ uniform draws **in expectation**. For another proposal generator, its internal random-number cost could differ.

::: {.callout-question title="In-Class Question"}

What goes wrong if we choose $M=1$ and simply replace acceptance probabilities above one by one? Why does a histogram that looks plausible fail to justify this shortcut?

:::

The envelope condition would fail near the mode. Capping the ratio changes the accepted density to one proportional to $\min\{g(y),f(y)/M\}$, which generally is not $f$.

### Unnormalized Densities and Statistical Connections

Sometimes we know only $h(x)$ where

$$
f(x)=\frac{h(x)}{Z},
\qquad Z=\int h(x)\,dx.
$$

If we can establish $h(x)\leq Cg(x)$ everywhere, we may accept with probability $h(Y)/\{Cg(Y)\}$ without knowing $Z$. The acceptance rate is then **$Z/C$**, not $1/C$.

This is useful for some posterior distributions, where the normalizing constant is difficult to compute. The challenge remains finding an easy proposal and a valid global bound.

**Connection to optimization.** Finding the smallest bound involves maximizing $f/g$, or its logarithm. A numerical optimizer or a finite grid may help explore this ratio, but does not by itself prove a global bound, especially on unbounded support.

**Connection to importance sampling.** Both methods use a proposal $g$ and the ratio $f/g$. Rejection sampling discards proposals to obtain draws from the target; importance sampling keeps proposals and weights their contributions when estimating an expectation.

**Computational limitation.** In high dimensions, a proposal that misses the target's concentration or dependence can yield a very small acceptance rate. This helps motivate more specialized methods, including the Markov chain Monte Carlo methods encountered in other statistics courses.

## Using Probability Distribution Theory

Known distributional relationships are also simulation algorithms. They often reveal *why* a distribution appears in statistical inference.

### Normal Variables from Uniforms: Box-Muller

If $U_1,U_2$ are independent uniforms, define

$$
R=\sqrt{-2\log U_1},\qquad \Theta=2\pi U_2,
$$

and

$$
Z_1=R\cos\Theta,\qquad Z_2=R\sin\Theta.
$$

Then $Z_1$ and $Z_2$ are independent standard normal variables.

**Geometric intuition.** A bivariate standard normal has circularly symmetric density. Its direction is uniform, and its squared radius has a $\chi^2_2$ distribution, equivalently an exponential distribution with rate $1/2$. Box-Muller generates that radius and direction, then converts polar coordinates to Cartesian coordinates. The polar Jacobian contributes a factor $r$, giving the radial density $r e^{-r^2/2}$.

```{r}
#| label: fig-rv-box-muller
#| code-fold: true
#| fig-cap: "Box-Muller creates a circular normal cloud from two independent uniforms. The two normal coordinates are independent even though their formulas share a radius and angle."
set.seed(8680)
n <- 4000
radius <- sqrt(-2 * log(runif(n)))
angle <- 2 * pi * runif(n)
z1 <- radius * cos(angle)
z2 <- radius * sin(angle)

ggplot(data.frame(z1, z2), aes(z1, z2)) +
  geom_point(alpha = 0.2, color = "steelblue", size = 1) +
  coord_equal() +
  labs(x = expression(Z[1]), y = expression(Z[2]))
```

Shared ingredients do not automatically imply dependence, and an approximately zero sample correlation does not prove independence. Here independence follows from the joint density. For routine work, use `rnorm()`; this construction explains one way a normal generator can work.

### Chi-Square, Student's t, and F

If $Z_1,\ldots,Z_k$ are independent $N(0,1)$ variables,

$$
V=\sum_{j=1}^k Z_j^2\sim\chi^2_k.
$$

If $Z\sim N(0,1)$ is independent of $V\sim\chi^2_\nu$, then

$$
T=\frac{Z}{\sqrt{V/\nu}}\sim t_\nu.
$$

If $V_1\sim\chi^2_{d_1}$ and $V_2\sim\chi^2_{d_2}$ are independent, then

$$
\frac{V_1/d_1}{V_2/d_2}\sim F_{d_1,d_2}.
$$

**Why this matters in inference.** A $t$ statistic divides a standardized normal estimation error by an estimated scale. When the scale estimate is unusually small, the ratio can be large. This explains the heavier tails of the $t$ distribution and the larger critical values used when estimating an unknown normal variance.

```{r}
#| label: fig-rv-student-t
#| code-fold: true
#| fig-cap: "A random denominator creates heavier tails. The generated ratios agree with the t distribution with 5 degrees of freedom; the dashed normal density has lighter tails."
set.seed(8681)
n <- 10000
nu <- 5
normal_numerator <- rnorm(n)
chi_square <- rowSums(matrix(rnorm(n * nu), nrow = n)^2)
t_draws <- normal_numerator / sqrt(chi_square / nu)

ggplot(data.frame(value = t_draws), aes(value)) +
  geom_histogram(aes(y = after_stat(density)), binwidth = 0.2,
                 fill = "lightblue", color = "white") +
  stat_function(fun = function(x) dt(x, df = nu),
                color = "firebrick", linewidth = 1) +
  stat_function(fun = dnorm, linetype = "dashed", linewidth = 0.8) +
  coord_cartesian(xlim = c(-6, 6)) +
  labs(x = "Ratio", y = "Density")
```

The numerator and denominator must be generated independently. Reusing the same normals in both changes the distribution. In this example, $\operatorname{Var}(T)=\nu/(\nu-2)=5/3$, so the raw $t_5$ distribution also has greater variance than $N(0,1)$.

### Beta Variables from Gamma Variables

If $G_1\sim\operatorname{Gamma}(a,\lambda)$ and $G_2\sim\operatorname{Gamma}(b,\lambda)$ are independent with the **same rate**, then

$$
X=\frac{G_1}{G_1+G_2}\sim\operatorname{Beta}(a,b).
$$

The ratio is a random share of a positive total. This gives intuition for beta models of probabilities and proportions. The common rate cancels from the ratio.

```{r}
#| label: fig-rv-beta-qq
#| code-fold: true
#| fig-cap: "A quantile comparison checks the entire distribution of the gamma ratio against Beta(3,2), including its tails."
set.seed(8682)
n <- 5000
g1 <- rgamma(n, shape = 3, rate = 1)
g2 <- rgamma(n, shape = 2, rate = 1)
beta_ratio <- g1 / (g1 + g2)

qq_data <- data.frame(
  theoretical = qbeta(ppoints(n), shape1 = 3, shape2 = 2),
  observed = sort(beta_ratio)
)
ggplot(qq_data, aes(theoretical, observed)) +
  geom_point(alpha = 0.3, size = 0.8, color = "steelblue") +
  geom_abline(intercept = 0, slope = 1, color = "firebrick") +
  coord_equal() +
  labs(x = "Theoretical Beta(3,2) quantile", y = "Sample quantile")
```

Normalizing several independent gammas with a common rate similarly produces a **Dirichlet** random vector. Its components sum to one, making it useful for uncertain category probabilities and mixture weights.

## Sums, Mixtures, and Hierarchical Models

### Sums and Convolution

A sum combines contributions from **all** its terms:

$$
S=X_1+\cdots+X_k.
$$

For independent continuous variables, the density of $X_1+X_2$ is

$$
f_S(s)=\int f_{X_1}(x)f_{X_2}(s-x)\,dx.
$$

The integral is a convolution. Simulation avoids computing this integral: generate the terms jointly according to the model and add them.

Some familiar relationships are:

- A sum of independent $\chi^2$ variables is chi-square with the sum of the degrees of freedom.
- A sum of $k$ independent $\operatorname{Exp}(\lambda)$ variables is $\operatorname{Gamma}(k,\lambda)$.
- A sum of $k$ independent geometric failure counts with success probability $p$ is negative binomial with `size = k` and `prob = p`.

**Practical intuition.** A total processing time adds durations from all stages. A count of failures before the $k$th success adds the failures before each successive success. The sum corresponds to an actual mechanism, not just an algebraic identity.

### Mixtures: Select a Component, Then Generate an Observation

A finite mixture has CDF and density

$$
F_X(x)=\sum_{j=1}^K w_jF_j(x),
\qquad
f_X(x)=\sum_{j=1}^K w_jf_j(x),
$$

where $w_j\geq0$ and $\sum_jw_j=1$.

Introduce an unobserved label $C$:

$$
P(C=j)=w_j,\qquad X\mid C=j\sim F_j.
$$

The law of total probability gives the mixture distribution after we average over the unknown label.

::: {.callout-algorithm title="Generating a Finite Mixture"}

For every observation:

1. Draw a component label $C$ with probabilities $w_1,\ldots,w_K$.
2. Draw one observation from $F_C$.

:::

The weights average **distributions**. They do not instruct us to return the numerical weighted average $w_1X_1+\cdots+w_KX_K$.

::: {.callout-example title="Three Different Data-Generating Mechanisms"}

Let $X_1\sim N(0,1)$ and $X_2\sim N(4,1)$ be independent. Compare:

$$
\begin{array}{ll}
\text{Sum:} & S=X_1+X_2\sim N(4,2),\\
\text{Weighted average:} & A=(X_1+X_2)/2\sim N(2,1/2),\\
\text{Mixture:} & X\sim 0.5N(0,1)+0.5N(4,1).
\end{array}
$$

The mixture has mean 2 and variance 5. It describes observations from two subpopulations; each observation belongs to one subpopulation.

```{r}
#| label: fig-rv-mixture
#| code-fold: true
#| fig-cap: "Adding two observations, averaging them, and selecting one population produce different distributions. The solid curves show their respective theoretical densities."
#| fig-width: 10
set.seed(8683)
n <- 10000
x1 <- rnorm(n, mean = 0, sd = 1)
x2 <- rnorm(n, mean = 4, sd = 1)
component <- sample.int(2, n, replace = TRUE, prob = c(0.5, 0.5))
x_mixture <- rnorm(n, mean = c(0, 4)[component], sd = 1)

mechanisms <- c("1. Sum", "2. Weighted average", "3. Mixture")
mixture_plot <- data.frame(
  value = c(x1 + x2, (x1 + x2) / 2, x_mixture),
  mechanism = rep(mechanisms, each = n)
)
grid <- seq(-4, 10, length.out = 500)
mixture_curves <- data.frame(
  value = rep(grid, 3),
  density = c(
    dnorm(grid, 4, sqrt(2)),
    dnorm(grid, 2, sqrt(0.5)),
    0.5 * dnorm(grid, 0, 1) + 0.5 * dnorm(grid, 4, 1)
  ),
  mechanism = rep(mechanisms, each = length(grid))
)
ggplot(mixture_plot, aes(value)) +
  geom_histogram(aes(y = after_stat(density)), bins = 45,
                 fill = "lightblue", color = "white") +
  geom_line(data = mixture_curves, aes(y = density),
            color = "firebrick", linewidth = 0.9) +
  facet_wrap(~ mechanism, nrow = 1) +
  labs(x = "Value", y = "Density")
```

:::

If component $j$ has mean $\mu_j$ and variance $\sigma_j^2$, then

$$
\mathbb{E}(X)=\sum_jw_j\mu_j=\mu,
$$

and the law of total variance gives

$$
\begin{aligned}
\operatorname{Var}(X)
&=\mathbb{E}\{\operatorname{Var}(X\mid C)\}+\operatorname{Var}\{\mathbb{E}(X\mid C)\}\\
&=\underbrace{\sum_j w_j\sigma_j^2}_{\text{within-component variation}}
 +\underbrace{\sum_j w_j(\mu_j-\mu)^2}_{\text{between-component variation}}.
\end{aligned}
$$

This decomposition explains why unobserved heterogeneity can inflate variability. A mixture of distinct normal components is generally nonnormal, but it need not be bimodal. Identical components yield the original normal distribution.

**Connection to clustering and EM.** Simulation starts with the label and then generates the observation. In mixture-model estimation, we see the observation and infer uncertainty about its label. This reversal is the basis for the responsibility calculations in the EM algorithm.

::: {.callout-question title="In-Class Question"}

In the example above, the weighted average and the mixture both have mean 2. Why would checking only the sample mean fail to detect the wrong generator?

:::

### Continuous Mixing: Why Count Data Can Be Overdispersed

A mixture label need not take only finitely many values. Consider

$$
\Lambda\sim\operatorname{Gamma}(a,b),
\qquad
Y\mid\Lambda\sim\operatorname{Poisson}(\Lambda),
$$

where $b$ is a rate. Then

$$
\mathbb{E}(Y)=\frac{a}{b},
\qquad
\operatorname{Var}(Y)=\frac{a}{b}+\frac{a}{b^2}.
$$

The second term is variability in the underlying rate. Marginally, $Y$ is negative binomial with `size = a` and `prob = b / (b + 1)`.

```{r}
#| label: rv-poisson-gamma
#| code-fold: true
set.seed(8684)
n <- 20000
a <- 2
b <- 0.5
latent_rate <- rgamma(n, shape = a, rate = b)
counts <- rpois(n, lambda = latent_rate)

knitr::kable(data.frame(
  quantity = c("Mean", "Variance"),
  simulated = c(mean(counts), var(counts)),
  theoretical = c(a / b, a / b + a / b^2)
), digits = 3)
```

**Practical interpretation.** Customers, neighborhoods, or machines can have different event rates. Even when each unit follows a Poisson model conditional on its rate, the pooled counts may have variance greater than their mean. This is one reason negative binomial regression is useful.

We drew one independent rate for each count. If several observations share a single group-specific rate, they become dependent after that rate is integrated out. **Where we place a draw in the simulation loop determines which observations share uncertainty.**

## Generating Multivariate Random Variables

### Correct Marginals Do Not Guarantee Correct Dependence

Suppose two measurements each have a standard normal marginal distribution. We still need to specify how they move together. Independent `rnorm()` calls impose independence; they cannot represent a positive correlation merely because the marginal histograms look correct.

For a normal vector, the covariance matrix supplies this information:

$$
\boldsymbol X\sim N_d(\boldsymbol\mu,\boldsymbol\Sigma).
$$

If $\boldsymbol Z\sim N_d(\boldsymbol0,I_d)$ and $AA^\top=\boldsymbol\Sigma$, then

$$
\boldsymbol X=\boldsymbol\mu+A\boldsymbol Z.
$$

Indeed,

$$
\mathbb{E}(\boldsymbol X)=\boldsymbol\mu,\qquad
\operatorname{Cov}(\boldsymbol X)=AI_dA^\top=\boldsymbol\Sigma.
$$

The normality follows because every linear combination of $\boldsymbol X$ is a linear combination of independent normals.

### Cholesky Decomposition and R's Row Convention

R's `chol(Sigma)` returns an **upper triangular** matrix $R$ satisfying

$$
R^\top R=\boldsymbol\Sigma.
$$

If simulated observations are **rows** of an $n\times d$ matrix $Z$, use

$$
X=ZR+\boldsymbol1_n\boldsymbol\mu^\top.
$$

For a column vector, the corresponding factor is $A=R^\top$. Mixing up these orientations generally produces the wrong covariance. See the [R documentation for `chol()`](https://stat.ethz.ch/R-manual/R-devel/library/base/html/chol.html).

```{r}
#| label: rv-mvn-generator
#| code-fold: true
rmvn_chol <- function(n, mu, Sigma) {
  d <- length(mu)
  stopifnot(
    length(n) == 1L, is.finite(n), n >= 0, n == floor(n),
    d > 0L, all(is.finite(mu)),
    is.matrix(Sigma), all(dim(Sigma) == c(d, d)),
    all(is.finite(Sigma)), isSymmetric(Sigma)
  )
  R <- chol(Sigma)  # requires positive definiteness here
  Z <- matrix(rnorm(n * d), nrow = n, ncol = d)
  sweep(Z %*% R, MARGIN = 2, STATS = mu, FUN = "+")
}

set.seed(8685)
mu <- c(1, 2)
Sigma <- matrix(c(1, 0.8, 0.8, 2), nrow = 2)
mvn_draws <- rmvn_chol(10000, mu, Sigma)

colMeans(mvn_draws)
cov(mvn_draws)
Sigma
```

The function also returns an $n\times d$ matrix for $n=0$ or $n=1$. This matters when a component in a simulated mixture receives few or no observations.

### Geometry, PCA, and Nearly Singular Covariance

The spectral decomposition is

$$
\boldsymbol\Sigma=P\Lambda P^\top,
\qquad
\boldsymbol\Sigma^{1/2}=P\Lambda^{1/2}P^\top.
$$

The eigenvectors specify directions, and the square roots of the eigenvalues specify how much to stretch a standard normal cloud in those directions.

This connects generation to **principal component analysis**. PCA identifies directions of variation in observed data; simulation constructs that variation from independent coordinates.

A covariance matrix can be positive semidefinite with some eigenvalues equal to zero. The normal vector then lies in a lower-dimensional subspace. Ordinary unpivoted Cholesky may fail, whereas an eigenvalue construction can handle that case. Tiny negative eigenvalues caused by roundoff may be set to zero after a tolerance check; substantial negative eigenvalues indicate an invalid covariance matrix and should not be silently removed.

If $\boldsymbol\Sigma$ stays fixed across many simulation batches, compute its factor once and reuse it. Factorization costs roughly order $d^3$, while multiplying $n$ observations by a dense factor costs order $nd^2$. Avoid repeating work that does not change the model.

### Dependence Changes Joint Probabilities

::: {.callout-example title="Two Measurements Exceed a Threshold"}

Consider two standardized laboratory measurements with unit marginal variances. We compare independence with correlation $\rho=0.8$ while keeping the marginals the same.

```{r}
#| label: fig-rv-dependence
#| code-fold: true
#| fig-cap: "The marginal distributions are the same in both panels. Positive correlation puts more probability in the upper-right region where both measurements exceed 1."
#| fig-width: 9
set.seed(8686)
n <- 30000
independent_pair <- rmvn_chol(n, c(0, 0), diag(2))
correlated_pair <- rmvn_chol(
  n, c(0, 0), matrix(c(1, 0.8, 0.8, 1), 2)
)

joint_probability <- c(
  mean(independent_pair[, 1] > 1 & independent_pair[, 2] > 1),
  mean(correlated_pair[, 1] > 1 & correlated_pair[, 2] > 1)
)
joint_results <- data.frame(
  model = c("Independent", "Correlation 0.8"),
  probability = joint_probability,
  MCSE = sqrt(joint_probability * (1 - joint_probability) / n)
)

# Plot a subset for legibility; estimates use all n observations.
dependence_plot <- rbind(
  data.frame(x = independent_pair[1:2000, 1],
             y = independent_pair[1:2000, 2], model = "Independent"),
  data.frame(x = correlated_pair[1:2000, 1],
             y = correlated_pair[1:2000, 2], model = "Correlation 0.8")
)
ggplot(dependence_plot, aes(x, y)) +
  annotate("rect", xmin = 1, xmax = Inf, ymin = 1, ymax = Inf,
           fill = "gold", alpha = 0.15) +
  geom_point(alpha = 0.2, color = "steelblue", size = 0.8) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_hline(yintercept = 1, linetype = "dashed") +
  facet_wrap(~ model) +
  coord_equal() +
  labs(x = "Measurement 1", y = "Measurement 2")
```

```{r}
#| label: rv-joint-probability
#| code-fold: true
knitr::kable(joint_results, digits = 4,
             col.names = c("Model", "Joint probability", "MCSE"))
(1 - pnorm(1))^2  # exact probability under independence
```

The simulated joint probability rises from approximately `r round(joint_probability[1], 3)` to `r round(joint_probability[2], 3)` under these two models. The **Monte Carlo standard error (MCSE)** measures simulation noise; we derive it below.

The same issue arises with simultaneous equipment failures, correlated regression errors, and repeated measurements. A study of a joint event must preserve the relevant dependence.

:::

### Multivariate Normal Mixtures

Each observation needs **one component label for the entire vector**. Drawing a different label for each coordinate changes the joint model.

```{r}
#| label: fig-rv-mvn-mixture
#| code-fold: true
#| fig-cap: "Each point is generated from one bivariate normal component. The label determines both its coordinates and its covariance structure."
set.seed(8687)
n <- 3000
component <- sample.int(2, n, replace = TRUE, prob = c(0.65, 0.35))
means <- list(c(0, 0), c(3, 2))
covariances <- list(
  matrix(c(1, 0.6, 0.6, 1), 2),
  matrix(c(0.6, -0.3, -0.3, 1), 2)
)
mixed_vectors <- matrix(NA_real_, nrow = n, ncol = 2)

for (j in seq_len(2)) {
  rows <- which(component == j)
  if (length(rows) > 0) {
    mixed_vectors[rows, ] <- rmvn_chol(
      length(rows), means[[j]], covariances[[j]]
    )
  }
}

ggplot(
  data.frame(x = mixed_vectors[, 1], y = mixed_vectors[, 2],
             component = factor(component)),
  aes(x, y, color = component)
) +
  geom_point(alpha = 0.25, size = 1) +
  coord_equal() +
  scale_color_manual(values = c("steelblue", "firebrick")) +
  labs(x = "Variable 1", y = "Variable 2", color = "Component")
```

For vector means $\boldsymbol\mu_j$ and covariances $\boldsymbol\Sigma_j$,

$$
\operatorname{Cov}(\boldsymbol X)=
\sum_jw_j\boldsymbol\Sigma_j+
\sum_jw_j(\boldsymbol\mu_j-\boldsymbol\mu)
        (\boldsymbol\mu_j-\boldsymbol\mu)^\top.
$$

Between-group differences can create marginal association even if variables are independent *within* every group. This is useful when designing simulation studies for clustering or confounding.

### Optional: Sample Covariance and the Wishart Distribution

If $\boldsymbol X_1,\ldots,\boldsymbol X_m$ are independent $N_d(\boldsymbol0,\boldsymbol\Sigma)$ vectors, then

$$
W=\sum_{i=1}^m\boldsymbol X_i\boldsymbol X_i^\top
\sim\operatorname{Wishart}_d(m,\boldsymbol\Sigma).
$$

With observations stored in rows, this is simply `crossprod(X)`. For a normal sample of size $n$ with an estimated mean, its usual sample covariance satisfies

$$
(n-1)S\sim\operatorname{Wishart}_d(n-1,\boldsymbol\Sigma).
$$

This explains why Wishart simulation is relevant to uncertainty in covariance estimates and multivariate inference.

```{r}
#| label: rv-wishart
#| code-fold: true
set.seed(8688)
Sigma_w <- matrix(c(1, 0.4, 0.4, 2), 2)
wishart_df <- 12
W <- crossprod(rmvn_chol(wishart_df, c(0, 0), Sigma_w))
W / wishart_df  # expectation equals Sigma_w, but one draw is noisy
```

## Uniform Directions and Uniform Points in a Ball

### Uniform on the Surface of a Sphere

Let $\boldsymbol Z\sim N_d(\boldsymbol0,I_d)$. Its distribution is unchanged by rotations, so it has no preferred direction. Consequently,

$$
\boldsymbol U=\frac{\boldsymbol Z}{\|\boldsymbol Z\|}
$$

is uniform on the **surface** of the unit sphere in $\mathbb R^d$. The zero vector has probability zero under the ideal normal model.

```{r}
#| label: rv-sphere-generator
#| code-fold: true
runif_sphere <- function(n, d) {
  stopifnot(
    length(n) == 1L, is.finite(n), n >= 0, n == floor(n),
    length(d) == 1L, is.finite(d), d >= 1, d == floor(d)
  )
  Z <- matrix(rnorm(n * d), nrow = n, ncol = d)
  lengths <- sqrt(rowSums(Z^2))
  sweep(Z, MARGIN = 1, STATS = lengths, FUN = "/")
}
```

We divide each row by its length directly. Constructing an $n\times n$ diagonal matrix of inverse lengths would waste memory.

### Uniform Inside the Ball

Uniform direction alone puts every point at distance 1. To fill the ball uniformly in volume, independently generate a radius $R$ satisfying

$$
P(R\leq r)=r^d,\quad 0\leq r\leq1.
$$

This follows because a ball of radius $r$ has $r^d$ times the volume of the unit ball. Therefore,

$$
R=V^{1/d},\qquad \boldsymbol X=R\boldsymbol U,
$$

where $V$ is uniform and independent of the direction.

```{r}
#| label: fig-rv-ball
#| code-fold: true
#| fig-cap: "In two dimensions, normalized normals lie on the circle. Multiplying by an independent square-root-uniform radius fills the disk uniformly in area."
#| fig-width: 9
set.seed(8689)
n <- 2500
directions <- runif_sphere(n, d = 2)
disk <- sweep(directions, 1, sqrt(runif(n)), FUN = "*")

sphere_plot <- rbind(
  data.frame(x = directions[, 1], y = directions[, 2],
             region = "Surface: unit circle"),
  data.frame(x = disk[, 1], y = disk[, 2],
             region = "Interior: unit disk")
)
ggplot(sphere_plot, aes(x, y)) +
  geom_point(alpha = 0.3, size = 0.7, color = "steelblue") +
  facet_wrap(~ region) +
  coord_equal() +
  labs(x = "Coordinate 1", y = "Coordinate 2")
```

**Why this matters.** Random directions appear in rotationally invariant perturbations, geometric Monte Carlo methods, and randomized algorithms. In high dimensions, a uniform point in a ball is typically close to its boundary: $P(R\leq0.9)=0.9^d$. At $d=100$, this probability is about $0.000027$.

::: {.callout-question title="In-Class Question"}

Why does a uniform radius on $(0,1)$ place too much mass near the center of a disk? Why does choosing a uniform latitude fail to generate uniform points on the surface of a three-dimensional sphere?

:::

## How Do We Know a Generator Is Working?

### Check More Than a Histogram

A histogram is a useful first view, but its appearance depends on bin width and sample size. Different distributions can have the same mean, variance, or a similar-looking central region.

A convincing validation combines theory with checks that target different possible mistakes:

| Check | What it can reveal |
|---|---|
| Support and constraints | Negative waiting times, noninteger counts, proportions outside $(0,1)$ |
| Mean and variance | Incorrect location, scale, or parameterization |
| Quantiles, ECDF, and tail probabilities | Distributional errors hidden by matching moments |
| Correlations and joint events | Incorrect dependence or matrix orientation |
| Repeated runs and random-number use | Reused seeds or repeated observations |
| Acceptance rate and proposal count | A costly envelope or an implementation error |

For the rejection sampler, the ECDF should track the target CDF, and the transformed values should be approximately uniform.

```{r}
#| label: fig-rv-validation
#| code-fold: true
#| fig-cap: "The empirical CDF of accepted proposals tracks the Beta(2,2) CDF. Sampling variation creates small discrepancies even for a correct generator."
ggplot(data.frame(value = beta_tight$draws), aes(value)) +
  stat_ecdf(geom = "step", color = "steelblue", linewidth = 0.7) +
  stat_function(fun = function(x) pbeta(x, 2, 2),
                color = "firebrick", linewidth = 0.9, linetype = "dashed") +
  labs(x = "Accepted value", y = "Cumulative probability")
```

Formal goodness-of-fit tests can supplement these checks, but passing a test is not a proof of correctness. A correct generator can occasionally be rejected, and a small sample may fail to detect a faulty one. The algorithm's mathematical justification remains essential.

### Monte Carlo Error

Suppose we estimate $p=P(X>c)$ using $B$ independent simulated observations:

$$
\hat p=\frac{1}{B}\sum_{b=1}^B\mathbf1\{X_b>c\}.
$$

Each indicator is Bernoulli, so

$$
\operatorname{Var}(\hat p)=\frac{p(1-p)}{B},
\qquad
\widehat{\operatorname{MCSE}}(\hat p)
=\sqrt{\frac{\hat p(1-\hat p)}{B}}.
$$

For an estimated mean of independent $h(X_b)$ with finite variance, the corresponding MCSE is

$$
\widehat{\operatorname{MCSE}}
\left(\frac1B\sum_{b=1}^B h(X_b)\right)
=\frac{s_h}{\sqrt B}.
$$

```{r}
#| label: rv-mcse
#| code-fold: true
set.seed(8690)
B <- 10000
event <- rexp(B, rate = 0.7) > 3
p_hat <- mean(event)
mcse <- sqrt(p_hat * (1 - p_hat) / B)

knitr::kable(data.frame(
  estimate = p_hat,
  MCSE = mcse,
  exact = exp(-0.7 * 3),
  lower_MC_limit = p_hat - 1.96 * mcse,
  upper_MC_limit = p_hat + 1.96 * mcse
), digits = 4,
col.names = c("Estimate", "MCSE", "Exact", "Lower MC limit", "Upper MC limit"))
```

These limits quantify numerical uncertainty from the finite simulation. They do not quantify uncertainty about whether the exponential model or its rate describes a real population.

To halve MCSE, we generally need about **four times** as many independent replications. For rare events, zero simulated occurrences do not imply zero probability: the plug-in MCSE also becomes zero and is misleading. A binomial interval or a method designed for rare events is then more informative.

::: {.callout-note title="Three Different Sources of Uncertainty"}

**Sampling uncertainty** concerns what would happen with another real dataset. **Monte Carlo uncertainty** comes from using finitely many simulated replications to study that behavior. **Model uncertainty** concerns whether the assumed mechanism is an adequate description of the setting.

More simulation reduces Monte Carlo uncertainty. It does not automatically reduce the other two.

:::

## Real-World Applications

The following examples use illustrative models and parameters. Each example states a decision-relevant quantity, translates the model into a generator, and checks the resulting simulation.

### Experimental Design: Estimating Power

::: {.callout-example title="How Often Would an Experiment Detect an Effect?"}

Suppose an experiment compares two independent groups, each with $n$ observations:

$$
X_i\sim N(0,\sigma^2),
\qquad
Y_i\sim N(\delta,\sigma^2).
$$

Here $\delta$ is a meaningful difference we want to detect. To keep the benchmark exact, assume $\sigma=1$ is known. The two-sided level-$0.05$ test rejects when

$$
|Z|>z_{0.975},
\qquad
Z=\frac{\bar Y-\bar X}{\sigma\sqrt{2/n}}.
$$

For $\delta=0$, the rejection probability is the Type I error rate. For $\delta\neq0$, it is power.

We generate **whole datasets**, apply the proposed analysis to each dataset, and record the result. The data size $n$ determines how informative each experiment is. The number of replications $B$ determines how accurately we estimate its rejection probability.

```{r}
#| label: rv-power-study
#| code-fold: true
set.seed(8691)
B <- 4000
sigma <- 1
design <- expand.grid(n = c(10, 20, 40, 80), delta = c(0, 0.5))
critical <- qnorm(0.975)

power_results <- do.call(rbind, lapply(seq_len(nrow(design)), function(i) {
  n <- design$n[i]
  delta <- design$delta[i]
  control <- matrix(rnorm(B * n, 0, sigma), nrow = B)
  treatment <- matrix(rnorm(B * n, delta, sigma), nrow = B)
  se <- sigma * sqrt(2 / n)
  statistic <- (rowMeans(treatment) - rowMeans(control)) / se
  estimated <- mean(abs(statistic) > critical)
  shift <- delta / se
  exact <- pnorm(-critical - shift) +
    pnorm(critical - shift, lower.tail = FALSE)

  data.frame(
    n = n, delta = delta, estimated = estimated, exact = exact,
    MCSE = sqrt(estimated * (1 - estimated) / B)
  )
}))
power_results$scenario <- ifelse(
  power_results$delta == 0, "Null: Type I error", "Effect 0.5: power"
)
knitr::kable(power_results[, c("n", "delta", "estimated", "exact", "MCSE")],
             digits = 4)
```

```{r}
#| label: fig-rv-power
#| code-fold: true
#| fig-cap: "Points are simulated rejection rates with approximate 95% Monte Carlo error bars. Lines are exact normal-model probabilities. Increasing the experiment's sample size increases power, while the Type I error remains near 0.05."
#| fig-width: 9
ggplot(power_results, aes(n, estimated)) +
  geom_line(aes(y = exact), color = "firebrick", linewidth = 0.9) +
  geom_errorbar(aes(
    ymin = pmax(0, estimated - 1.96 * MCSE),
    ymax = pmin(1, estimated + 1.96 * MCSE)
  ), width = 3, color = "steelblue") +
  geom_point(color = "steelblue", size = 2) +
  facet_wrap(~ scenario) +
  scale_x_continuous(breaks = c(10, 20, 40, 80)) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(x = "Sample size per group", y = "Rejection probability")
```

**Interpretation.** The exact formula gives a benchmark for our simulation. When the model becomes more realistic, such as unequal variances, clustered observations, or missing outcomes, simulation may remain feasible even when a simple power formula does not.

If $\sigma$ is estimated, replace the known-variance test with the actual planned procedure, such as a two-sample $t$ test. Simulating one analysis and reporting power for another answers the wrong question.

:::

### Quality Control: Predicting Future Defects with an Uncertain Rate

::: {.callout-example title="A Predictive Distribution Needs More Than a Plug-In Estimate"}

Suppose a process produces independent binary defect indicators conditional on an unknown defect probability $p$. Start with a $\operatorname{Beta}(1,1)$ prior and observe 3 defects among 18 items. Beta-binomial conjugacy gives

$$
p\mid\text{data}\sim\operatorname{Beta}(4,16).
$$

To predict the number of defects $Y_{\mathrm{new}}$ in a future batch of $m=50$ items, generate

$$
p^{(b)}\sim\operatorname{Beta}(4,16),
\qquad
Y_{\mathrm{new}}^{(b)}\mid p^{(b)}
\sim\operatorname{Binomial}(50,p^{(b)}).
$$

Each simulated future batch uses one shared draw of its possible process rate. This integrates uncertainty about $p$ into the prediction.

The plug-in alternative uses only $\bar p=4/20=0.2$:

$$
Y_{\mathrm{plug-in}}\sim\operatorname{Binomial}(50,0.2).
$$

Both have mean 10, but the posterior predictive variance is larger:

$$
\begin{aligned}
\operatorname{Var}(Y_{\mathrm{new}}\mid\text{data})
&=m\bar p(1-\bar p)\frac{a+b+m}{a+b+1}\\
&=50(0.2)(0.8)\frac{70}{21}
\approx26.67,
\end{aligned}
$$

compared with the plug-in variance $50(0.2)(0.8)=8$.

```{r}
#| label: rv-predictive
#| code-fold: true
set.seed(8692)
B <- 20000
a_post <- 4
b_post <- 16
batch_size <- 50
p_mean <- a_post / (a_post + b_post)

p_draw <- rbeta(B, shape1 = a_post, shape2 = b_post)
predictive_defects <- rbinom(B, size = batch_size, prob = p_draw)
plugin_defects <- rbinom(B, size = batch_size, prob = p_mean)

k <- 0:batch_size
# Exact beta-binomial probabilities, computed on the log scale.
predictive_pmf <- exp(
  lchoose(batch_size, k) +
    lbeta(k + a_post, batch_size - k + b_post) -
    lbeta(a_post, b_post)
)
predictive_comparison <- data.frame(
  model = c("Posterior predictive", "Plug-in"),
  simulated_variance = c(var(predictive_defects), var(plugin_defects)),
  theoretical_variance = c(
    batch_size * p_mean * (1 - p_mean) *
      (a_post + b_post + batch_size) / (a_post + b_post + 1),
    batch_size * p_mean * (1 - p_mean)
  ),
  simulated_prob_at_least_15 = c(
    mean(predictive_defects >= 15), mean(plugin_defects >= 15)
  ),
  exact_prob_at_least_15 = c(
    sum(predictive_pmf[k >= 15]),
    pbinom(14, size = batch_size, prob = p_mean, lower.tail = FALSE)
  )
)
knitr::kable(predictive_comparison, digits = 3, col.names = c(
  "Model", "Sim. variance", "Exact variance", "Sim. P(Y >= 15)", "Exact P(Y >= 15)"
))
```

```{r}
#| label: fig-rv-predictive
#| code-fold: true
#| fig-cap: "The posterior predictive distribution spreads probability over a wider range of defect counts. The plug-in model ignores uncertainty about the process rate. Lines show exact probabilities and bars show simulated proportions."
#| fig-width: 9
predictive_plot <- rbind(
  data.frame(
    defects = k,
    proportion = tabulate(predictive_defects + 1, nbins = batch_size + 1) / B,
    exact = predictive_pmf,
    model = "Posterior predictive"
  ),
  data.frame(
    defects = k,
    proportion = tabulate(plugin_defects + 1, nbins = batch_size + 1) / B,
    exact = dbinom(k, size = batch_size, prob = p_mean),
    model = "Plug-in"
  )
)
ggplot(predictive_plot, aes(defects, proportion)) +
  geom_col(fill = "lightblue", width = 0.85) +
  geom_line(aes(y = exact), color = "firebrick", linewidth = 0.8) +
  geom_vline(xintercept = 14.5, linetype = "dashed") +
  facet_wrap(~ model) +
  coord_cartesian(xlim = c(0, 35)) +
  labs(x = "Defects in the next 50 items", y = "Probability")
```

**Why it matters.** A decision triggered by a large defect count depends on the predictive tail, not just the expected count. Hierarchical simulation carries parameter uncertainty through to that tail.

If we drew a fresh independent $p$ for each individual item instead of one shared $p$ for the batch, the marginal total would be binomial with probability $\mathbb{E}(p)$. That would represent a different mechanism and lose the shared uncertainty in this posterior prediction.

:::

### Service Operations: Random Arrivals Create Dependent Waiting Times

::: {.callout-example title="How Often Does a Customer Wait More Than Five Minutes?"}

Consider a single server, first-come-first-served, initially idle. Let interarrival times be independent exponentials with rate $\lambda$, and service times be independent $\operatorname{Gamma}(2,1)$ variables, with mean 2 minutes. Assume the arrival and service processes are independent.

If $A_i$ is the time between customers $i-1$ and $i$, $S_i$ is customer $i$'s service time, and $W_i$ is the wait before service, then

$$
W_1=0,\qquad
W_i=\max\{0,W_{i-1}+S_{i-1}-A_i\}.
$$

This recursion turns independent primitive random inputs into dependent waiting times. A long service can delay several later customers.

We compare arrival rates $0.35$ and $0.45$ per minute. Their utilization values are $\lambda\mathbb{E}(S)=0.7$ and $0.9$. Both are below one, but the system closer to capacity has less room to absorb bursts.

```{r}
#| label: rv-queue
#| code-fold: true
simulate_queue <- function(arrival_rate, n_customers = 200) {
  interarrival <- rexp(n_customers, rate = arrival_rate)
  service <- rgamma(n_customers, shape = 2, rate = 1)
  wait <- numeric(n_customers)
  if (n_customers > 1) {
    for (i in 2:n_customers) {
      wait[i] <- max(0, wait[i - 1] + service[i - 1] - interarrival[i])
    }
  }
  wait
}

set.seed(8693)
B_days <- 800
rates <- c(0.35, 0.45)
queue_results <- do.call(rbind, lapply(rates, function(rate) {
  # Each replication is an independent run of 200 customers, starting empty.
  fraction_delayed <- replicate(
    B_days, mean(simulate_queue(rate) > 5)
  )
  data.frame(
    arrival_rate = rate,
    utilization = 2 * rate,
    fraction_waiting_over_5 = mean(fraction_delayed),
    MCSE = sd(fraction_delayed) / sqrt(B_days)
  )
}))
knitr::kable(queue_results, digits = 4, col.names = c(
  "Arrival rate", "Utilization", "Fraction waiting > 5", "MCSE"
))
```

```{r}
#| label: fig-rv-queue
#| code-fold: true
#| fig-cap: "Two illustrative paths show how delays carry over between customers. Estimates in the table average 800 independent runs per scenario; one path alone can be misleading."
#| fig-width: 9
set.seed(8694)
queue_paths <- do.call(rbind, lapply(rates, function(rate) {
  data.frame(
    customer = 1:200, wait = simulate_queue(rate),
    scenario = paste("Arrival rate =", rate)
  )
}))
ggplot(queue_paths, aes(customer, wait)) +
  geom_line(color = "steelblue", linewidth = 0.6) +
  geom_hline(yintercept = 5, linetype = "dashed", color = "firebrick") +
  facet_wrap(~ scenario) +
  labs(x = "Customer index", y = "Wait before service (minutes)")
```

**Interpretation.** The estimand is the expected fraction of the first 200 customers who wait more than five minutes in a system starting empty. It is not automatically the long-run stationary probability. Studying steady-state performance requires addressing the initial transient and dependence.

We compute MCSE from **independent run-level fractions**. Pooling all customers and treating their wait indicators as independent Bernoulli trials would ignore the dependence created by the queue.

This example combines distribution choice, recursive computation, and statistical reasoning about the unit of replication.

:::

## Choosing a Generation Method

| Available structure | Natural approach | Main issue to check |
|---|---|---|
| A reliable built-in generator | Use the distribution's `r` function | Parameterization and joint assumptions |
| A tractable quantile | Inverse transformation | Correct generalized inverse and endpoints |
| A tractable CDF without a convenient inverse | Numerical root finding | Bracketing, monotonicity, tolerance, and cost |
| A target density and a suitable proposal | Acceptance-rejection | Global envelope and acceptance rate |
| A known distributional identity | Transform simpler independent variables | Conditions such as independence and common rates |
| A latent-class or hierarchical model | Generate latent variables, then observations | Which observations share each latent draw |
| A normal covariance structure | Matrix factorization | Positive definiteness and factor orientation |
| Rotational or volume symmetry | Geometric construction | Surface versus volume and the correct radius |

Methods are often combined. A hierarchical model may use inverse transformation for event times, normal matrix factors for random effects, and conditional Bernoulli draws for outcomes.

The goal is a generator that is statistically correct, numerically reliable, and efficient enough for the problem. A sophisticated algorithm cannot compensate for an incorrect data-generating mechanism.

## Questions to Think at Home

1. Derive an inverse-transform generator for $f(x)=4x^3$ on $(0,1)$. Check its theoretical mean and one tail probability.

2. Generate 500 observations from an exponential distribution with mean 5. Explain the difference between `rate = 5` and `rate = 1 / 5`.

3. Why does the generalized inverse use $F(x)\geq u$? Illustrate your answer with a Bernoulli distribution.

4. For a $\operatorname{Beta}(3,1)$ target and a uniform proposal, find the smallest valid rejection bound $M$, the acceptance probability, and the expected number of proposals for 1000 accepted draws.

5. Derive the acceptance probability when the target is known only through an unnormalized density. Which constant need not be computed?

6. Generate a $t_4$ variable from independent normal and chi-square variables. Explain why its variance is 2 rather than 1. How would you rescale it to unit variance?

7. Compare $0.3X_1+0.7X_2$ with a mixture that selects $X_1$ with probability $0.3$. Derive both means and variances for independent $X_1\sim N(0,1)$ and $X_2\sim N(4,1)$.

8. For a fixed positive definite covariance matrix, compare `Z %*% chol(Sigma)` with `Z %*% t(chol(Sigma))`. Derive the covariance produced by each.

9. In a mixture of bivariate normals, what changes if you draw one component label per coordinate instead of one per row?

10. Derive the radius distribution for a uniform point inside a three-dimensional ball. Verify it using the empirical distribution of simulated radii.

11. Extend the power experiment to unknown variance and a two-sample $t$ test. Compare Type I error and power under normal errors and under rescaled $t_3$ errors with the same variance.

12. Simulate repeated datasets and estimate the coverage of a 95% confidence interval for an exponential mean. State the true mean, interval procedure, number of datasets, and MCSE of the coverage estimate.

13. In the defect prediction example, vary the posterior concentration $a+b$ while holding $a/(a+b)=0.2$. What happens to predictive variance, and why?

14. In the queue example, why is the number of independent runs relevant to the MCSE calculation? How would the target quantity change if you analyzed only later customers after a warm-up period?

## Takeaways

- A probability model becomes operational when we can generate data from it.
- Uniform draws can be transformed into observations through a generalized inverse CDF.
- Rejection sampling reshapes an easy proposal; a valid, tight envelope controls efficiency.
- Distributional identities connect simulation to the foundations of statistical inference.
- A mixture selects a component; a sum combines contributions from all components.
- Hierarchical draws specify both marginal variation and dependence.
- Multivariate generation must reproduce covariance and joint behavior, not only marginal histograms.
- Validation combines mathematical reasoning, numerical checks, and distributional diagnostics.
- More independent replications reduce Monte Carlo error; they do not correct a wrong model.

The next step is to use these generators to approximate expectations, integrals, and sampling distributions. That is the central task of **Monte Carlo simulation**.

## References and Further Reading {.unnumbered}

This chapter adapts and expands the [2025 STAT 8670 chapter on generating random variables](https://chikuang.github.io/teaching/2025F_stat8670/03-generating-rv.html) and its [Quarto source in the master branch](https://github.com/chikuang/Teaching-Stat8670/blob/master/03-generating-rv.qmd).

For software details, consult the official R documentation linked at the relevant examples. Useful local help pages include `?RNGkind`, `?Distributions`, `?qgeom`, `?chol`, and `?rWishart`. The course's original reading is Rizzo, M. L. (2007), *Statistical Computing with R*, Chapman & Hall/CRC.
